---
title: "Statistical Rethinking (Code)"
author: "Chapter 5"
date: "February, 2017"
output:
  pdf_document: default
  html_document: default
params:
  original: no
  updated: yes
---

```{r, opts.label = "original", results = "asis", echo = FALSE}
cat("The original code from *Statistical Rethinking* is shown below.")
cat("\n\n")
```

```{r, opts.label = "updated", results = "asis", echo = FALSE}
cat(paste(
"Code from *Statistical Rethinking* modified by R Pruim is shown below.  Differences to the oringal include:",
"",
"  * a preference for putting data into containers (data frames, mostly), rather than working with lose vectors.",
"  * use of `lattice` and `ggplot2` rather than base graphics",
"  * use of `tidyverse` for data transformation",
"  * better (in my opinion) naming conventions",
sep = "\n"
))
cat("\n\n")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messae = FALSE)
require(mosaic)
require(rethinking)
```



## R code 5.1

```{r, chunk5.1}
# load data
library(rethinking)
data(WaffleDivorce)
d <- WaffleDivorce

# standardize predictor
d$MedianAgeMarriage.s <-
  (d$MedianAgeMarriage - mean(d$MedianAgeMarriage)) /
  sd(d$MedianAgeMarriage)

# fit model
m5.1 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bA * MedianAgeMarriage.s,
    a ~ dnorm(10, 10),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

```

## R code 5.2

```{r, chunk5.2}
# compute percentile interval of mean
MAM.seq <- seq(from = -3,
               to = 3.5,
               length.out = 30)
mu <- link(m5.1, data = data.frame(MedianAgeMarriage.s = MAM.seq))
mu.PI <- apply(mu, 2, PI)

# plot it all
plot(Divorce ~ MedianAgeMarriage.s, data = d, col = rangi2)
abline(m5.1)
shade(mu.PI, MAM.seq)

```

## R code 5.3

```{r, chunk5.3}
d$Marriage.s <- (d$Marriage - mean(d$Marriage)) / sd(d$Marriage)
m5.2 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bR * Marriage.s,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

```

## R code 5.4

```{r, chunk5.4}
m5.3 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bR * Marriage.s + bA * MedianAgeMarriage.s,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.3)

```

## R code 5.5

```{r, chunk5.5}
plot(precis(m5.3))

```

## R code 5.6

```{r, chunk5.6}
m5.4 <- map(
  alist(
    Marriage.s ~ dnorm(mu, sigma),
    mu <- a + b * MedianAgeMarriage.s,
    a ~ dnorm(0, 10),
    b ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

```

## R code 5.7

```{r, chunk5.7}
# compute expected value at MAP, for each State
mu <- coef(m5.4)['a'] + coef(m5.4)['b'] * d$MedianAgeMarriage.s
# compute residual for each State
m.resid <- d$Marriage.s - mu

```

## R code 5.8

```{r, chunk5.8}
plot(Marriage.s ~ MedianAgeMarriage.s, d, col = rangi2)
abline(m5.4)
# loop over States
for (i in 1:length(m.resid)) {
  x <- d$MedianAgeMarriage.s[i] # x location of line segment
  y <- d$Marriage.s[i] # observed endpoint of line segment
  # draw the line segment
  lines(c(x, x),
        c(mu[i], y),
        lwd = 0.5,
        col = col.alpha("black", 0.7))
}

```

## R code 5.9

```{r, chunk5.9}
# prepare new counterfactual data
A.avg <- mean(d$MedianAgeMarriage.s)
R.seq <- seq(from = -3,
             to = 3,
             length.out = 30)
pred.data <- data.frame(Marriage.s = R.seq,
                        MedianAgeMarriage.s = A.avg)

# compute counterfactual mean divorce (mu)
mu <- link(m5.3, data = pred.data)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# simulate counterfactual divorce outcomes
R.sim <- sim(m5.3, data = pred.data, n = 1e4)
R.PI <- apply(R.sim, 2, PI)

# display predictions, hiding raw data with type="n"
plot(Divorce ~ Marriage.s, data = d, type = "n")
mtext("MedianAgeMarriage.s = 0")
lines(R.seq, mu.mean)
shade(mu.PI, R.seq)
shade(R.PI, R.seq)

```

## R code 5.10

```{r, chunk5.10}
R.avg <- mean(d$Marriage.s)
A.seq <- seq(from = -3,
             to = 3.5,
             length.out = 30)
pred.data2 <- data.frame(Marriage.s = R.avg,
                         MedianAgeMarriage.s = A.seq)

mu <- link(m5.3, data = pred.data2)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

A.sim <- sim(m5.3, data = pred.data2, n = 1e4)
A.PI <- apply(A.sim, 2, PI)

plot(Divorce ~ MedianAgeMarriage.s,
     data = d,
     type = "n")
mtext("Marriage.s = 0")
lines(A.seq, mu.mean)
shade(mu.PI, A.seq)
shade(A.PI, A.seq)

```

## R code 5.11

```{r, chunk5.11}
# call link without specifying new data
# so it uses original data
mu <- link(m5.3)

# summarize samples across cases
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# simulate observations
# again no new data, so uses original data
divorce.sim <- sim(m5.3, n = 1e4)
divorce.PI <- apply(divorce.sim, 2, PI)

```

## R code 5.12

```{r, chunk5.12}
plot(
  mu.mean ~ d$Divorce,
  col = rangi2,
  ylim = range(mu.PI),
  xlab = "Observed divorce",
  ylab = "Predicted divorce"
)
abline(a = 0, b = 1, lty = 2)
for (i in 1:nrow(d))
  lines(rep(d$Divorce[i], 2), c(mu.PI[1, i], mu.PI[2, i]),
        col = rangi2)

```

## R code 5.13

```{r, chunk5.13, eval = FALSE}
identify(
  x = d$Divorce,
  y = mu.mean,
  labels = d$Loc,
  cex = 0.8
)

```

## R code 5.14

```{r, chunk5.14}
# compute residuals
divorce.resid <- d$Divorce - mu.mean
# get ordering by divorce rate
o <- order(divorce.resid)
# make the plot
dotchart(
  divorce.resid[o],
  labels = d$Loc[o],
  xlim = c(-6, 5),
  cex = 0.6
)
abline(v = 0, col = col.alpha("black", 0.2))
for (i in 1:nrow(d)) {
  j <- o[i] # which State in order
  lines(d$Divorce[j] - c(mu.PI[1, j], mu.PI[2, j]), rep(i, 2))
  points(
    d$Divorce[j] - c(divorce.PI[1, j], divorce.PI[2, j]),
    rep(i, 2),
    pch = 3,
    cex = 0.6,
    col = "gray"
  )
}

```

## R code 5.15

```{r, chunk5.15}
N <- 100                         # number of cases
x_real <-
  rnorm(N)             # x_real as Gaussian with mean 0 and stddev 1
x_spur <-
  rnorm(N, x_real)    # x_spur as Gaussian with mean=x_real
y <- rnorm(N, x_real)         # y as Gaussian with mean=x_real
d <- data.frame(y, x_real, x_spur) # bind all together in data frame

```

## R code 5.16

```{r, chunk5.16}
library(rethinking)
data(milk)
d <- milk
str(d)

```

## R code 5.17

```{r, chunk5.17, error = TRUE}
m5.5 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = d
)

```

## R code 5.18

```{r, chunk5.18}
d$neocortex.perc

```

## R code 5.19

```{r, chunk5.19}
dcc <- d[complete.cases(d),]

```

## R code 5.20

```{r, chunk5.20}
m5.5 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = dcc
)

```

## R code 5.21

```{r, chunk5.21}
precis(m5.5, digits = 3)

```

## R code 5.22

```{r, chunk5.22}
coef(m5.5)["bn"] * (76 - 55)

```

## R code 5.23

```{r, chunk5.23}
np.seq <- 0:100
pred.data <- data.frame(neocortex.perc = np.seq)

mu <- link(m5.5, data = pred.data, n = 1e4)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

plot(kcal.per.g ~ neocortex.perc, data = dcc, col = rangi2)
lines(np.seq, mu.mean)
lines(np.seq, mu.PI[1, ], lty = 2)
lines(np.seq, mu.PI[2, ], lty = 2)

```

## R code 5.24

```{r, chunk5.24}
dcc$log.mass <- log(dcc$mass)

```

## R code 5.25

```{r, chunk5.25}
m5.6 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bm * log.mass,
    a ~ dnorm(0, 100),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = dcc
)
precis(m5.6)

```

## R code 5.26

```{r, chunk5.26}
m5.7 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc + bm * log.mass,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = dcc
)
precis(m5.7)

```

## R code 5.27

```{r, chunk5.27}
mean.log.mass <- mean(log(dcc$mass))
np.seq <- 0:100
pred.data <- data.frame(neocortex.perc = np.seq,
                        log.mass = mean.log.mass)

mu <- link(m5.7, data = pred.data, n = 1e4)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

plot(kcal.per.g ~ neocortex.perc, data = dcc, type = "n")
lines(np.seq, mu.mean)
lines(np.seq, mu.PI[1, ], lty = 2)
lines(np.seq, mu.PI[2, ], lty = 2)

```

## R code 5.28

```{r, chunk5.28}
N <- 100                         # number of cases
rho <- 0.7                       # correlation btw x_pos and x_neg
x_pos <- rnorm(N)              # x_pos as Gaussian
x_neg <- rnorm(N, rho * x_pos,  # x_neg correlated with x_pos
               sqrt(1 - rho ^ 2))
y <-
  rnorm(N, x_pos - x_neg)  # y equally associated with x_pos, x_neg
d <- data.frame(y, x_pos, x_neg)   # bind all together in data frame

```

## R code 5.29

```{r, chunk5.29}
N <- 100                          # number of individuals
height <- rnorm(N, 10, 2)           # sim total height of each
leg_prop <- runif(N, 0.4, 0.5)      # leg as proportion of height
leg_left <-
  leg_prop * height +     # sim left leg as proportion + error
  rnorm(N, 0, 0.02)
leg_right <-
  leg_prop * height +    # sim right leg as proportion + error
  rnorm(N, 0, 0.02)
# combine into data frame
d <- data.frame(height, leg_left, leg_right)

```

## R code 5.30

```{r, chunk5.30}
m5.8 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * leg_left + br * leg_right,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.8)

```

## R code 5.31

```{r, chunk5.31}
plot(precis(m5.8))

```

## R code 5.32

```{r, chunk5.32}
post <- extract.samples(m5.8)
plot(bl ~ br, post, col = col.alpha(rangi2, 0.1), pch = 16)

```

## R code 5.33

```{r, chunk5.33}
sum_blbr <- post$bl + post$br
dens(sum_blbr,
     col = rangi2,
     lwd = 2,
     xlab = "sum of bl and br")

```

## R code 5.34

```{r, chunk5.34}
m5.9 <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- a + bl * leg_left,
  a ~ dnorm(10, 100),
  bl ~ dnorm(2, 10),
  sigma ~ dunif(0, 10)
),
data = d)
precis(m5.9)

```

## R code 5.35

```{r, chunk5.35}
library(rethinking)
data(milk)
d <- milk

```

## R code 5.36

```{r, chunk5.36}
# kcal.per.g regressed on perc.fat
m5.10 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bf * perc.fat,
    a ~ dnorm(0.6, 10),
    bf ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

# kcal.per.g regressed on perc.lactose
m5.11 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bl * perc.lactose,
    a ~ dnorm(0.6, 10),
    bl ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)

precis(m5.10, digits = 3)
precis(m5.11, digits = 3)

```

## R code 5.37

```{r, chunk5.37}
m5.12 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bf * perc.fat + bl * perc.lactose,
    a ~ dnorm(0.6, 10),
    bf ~ dnorm(0, 1),
    bl ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.12, digits = 3)

```

## R code 5.38

```{r, chunk5.38}
pairs(~ kcal.per.g + perc.fat + perc.lactose,
      data = d,
      col = rangi2)

```

## R code 5.39

```{r, chunk5.39}
cor(d$perc.fat, d$perc.lactose)

```

## R code 5.40

```{r, chunk5.40}
library(rethinking)
data(milk)
d <- milk
sim.coll <- function(r = 0.9) {
  d$x <- rnorm(nrow(d), mean = r * d$perc.fat,
               sd = sqrt((1 - r ^ 2) * var(d$perc.fat)))
  m <- lm(kcal.per.g ~ perc.fat + x, data = d)
  sqrt(diag(vcov(m)))[2] # stddev of parameter
}
rep.sim.coll <- function(r = 0.9, n = 100) {
  stddev <- replicate(n, sim.coll(r))
  mean(stddev)
}
r.seq <- seq(from = 0, to = 0.99, by = 0.01)
stddev <- sapply(r.seq, function(z)
  rep.sim.coll(r = z, n = 100))
plot(
  stddev ~ r.seq,
  type = "l",
  col = rangi2,
  lwd = 2,
  xlab = "correlation"
)

```

## R code 5.41

```{r, chunk5.41}
# number of plants
N <- 100

# simulate initial heights
h0 <- rnorm(N, 10, 2)

# assign treatments and simulate fungus and growth
treatment <- rep(0:1, each = N / 2)
fungus <- rbinom(N, size = 1, prob = 0.5 - treatment * 0.4)
h1 <- h0 + rnorm(N, 5 - 3 * fungus)

# compose a clean data frame
d <-
  data.frame(
    h0 = h0,
    h1 = h1,
    treatment = treatment,
    fungus = fungus
  )

```

## R code 5.42

```{r, chunk5.42}
m5.13 <- map(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- a + bh * h0 + bt * treatment + bf * fungus,
    a ~ dnorm(0, 100),
    c(bh, bt, bf) ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.13)

```

## R code 5.43

```{r, chunk5.43}
m5.14 <- map(
  alist(
    h1 ~ dnorm(mu, sigma),
    mu <- a + bh * h0 + bt * treatment,
    a ~ dnorm(0, 100),
    c(bh, bt) ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.14)

```

## R code 5.44

```{r, chunk5.44}
data(Howell1)
d <- Howell1
str(d)

```

## R code 5.45

```{r, chunk5.45}
m5.15 <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- a + bm * male,
  a ~ dnorm(178, 100),
  bm ~ dnorm(0, 10),
  sigma ~ dunif(0, 50)
),
data = d)
precis(m5.15)

```

## R code 5.46

```{r, chunk5.46}
post <- extract.samples(m5.15)
mu.male <- post$a + post$bm
PI(mu.male)

```

## R code 5.47

```{r, chunk5.47}
m5.15b <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- af * (1 - male) + am * male,
  af ~ dnorm(178, 100),
  am ~ dnorm(178, 100),
  sigma ~ dunif(0, 50)
),
data = d)

```

## R code 5.48

```{r, chunk5.48}
data(milk)
d <- milk
unique(d$clade)

```

## R code 5.49

```{r, chunk5.49}
(d$clade.NWM <- ifelse(d$clade == "New World Monkey", 1, 0))

```

## R code 5.50

```{r, chunk5.50}
d$clade.OWM <- ifelse(d$clade == "Old World Monkey", 1, 0)
d$clade.S <- ifelse(d$clade == "Strepsirrhine", 1, 0)

```

## R code 5.51

```{r, chunk5.51}
m5.16 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + b.NWM * clade.NWM + b.OWM * clade.OWM + b.S * clade.S,
    a ~ dnorm(0.6, 10),
    b.NWM ~ dnorm(0, 1),
    b.OWM ~ dnorm(0, 1),
    b.S ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = d
)
precis(m5.16)

```

## R code 5.52

```{r, chunk5.52}
# sample posterior
post <- extract.samples(m5.16)

# compute averages for each category
mu.ape <- post$a
mu.NWM <- post$a + post$b.NWM
mu.OWM <- post$a + post$b.OWM
mu.S <- post$a + post$b.S

# summarize using precis
precis(data.frame(mu.ape, mu.NWM, mu.OWM, mu.S))

```

## R code 5.53

```{r, chunk5.53}
diff.NWM.OWM <- mu.NWM - mu.OWM
quantile(diff.NWM.OWM, probs = c(0.025, 0.5, 0.975))

```

## R code 5.54

```{r, chunk5.54}
(d$clade_id <- coerce_index(d$clade))

```

## R code 5.55

```{r, chunk5.55}
m5.16_alt <- map(alist(
  kcal.per.g ~ dnorm(mu, sigma),
  mu <- a[clade_id],
  a[clade_id] ~ dnorm(0.6, 10),
  sigma ~ dunif(0, 10)
),
data = d)
precis(m5.16_alt, depth = 2)

```

## R code 5.56

```{r, chunk5.56}
m5.17 <- lm(y ~ 1 + x, data = d)
m5.18 <- lm(y ~ 1 + x + z + w, data = d)

```

## R code 5.57

```{r, chunk5.57}
m5.17 <- lm(y ~ 1 + x, data = d)
m5.19 <- lm(y ~ x, data = d)

```

## R code 5.58

```{r, chunk5.58}
m5.20 <- lm(y ~ 0 + x, data = d)
m5.21 <- lm(y ~ x - 1, data = d)

```

## R code 5.59

```{r, chunk5.59}
m5.22 <- lm(y ~ 1 + as.factor(season), data = d)

```

## R code 5.60

```{r, chunk5.60}
d$x2 <- d$x ^ 2
d$x3 <- d$x ^ 3
m5.23 <- lm(y ~ 1 + x + x2 + x3, data = d)

```

## R code 5.61

```{r, chunk5.61}
m5.24 <- lm(y ~ 1 + x + I(x ^ 2) + I(x ^ 3), data = d)

```

## R code 5.62

```{r, chunk5.62}
data(cars)
glimmer(dist ~ speed, data = cars)

```

