---
title: "Statistical Rethinking (Code)"
author: "Chapter 9"
date: "March, 2017"
output:
  html_document: 
    fig_height: 3.5
  pdf_document: 
    fig_height: 3.5
---

```{r, setup, include = FALSE}
# Load packages here 
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  cache = TRUE,
  tidy = FALSE,     # display code as typed
  size = "small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_light())
```

#### R Code 8.1

The book does this as a chunk of code.  It is much nicer to wrap this
in a function and to return a data frame that makes plotting nicer.

```{r, chunk8.1}
KingMarkov <- function(
  num_steps = 1e5,
  population = 1:10,
  start = 10) {
  
  num_islands <- length(population)
  position <- rep(NA, num_steps)  # trick to pre-alocate memory
  current <- start
  
  for (i in 1:num_steps) {
    # record current position
    position[i] <- current
    
    # flip coin to generate proposal
    proposal <- 1 + (current + sample(c(-2, 0), size = 1) ) %% 10
    
    # move?
    prob_move <- population[proposal] / population[current]
    current <- ifelse(runif(1) < prob_move, proposal, current)
  }
  data_frame(
    step = 1:num_steps,
    position = position
  )
}
```

In the short-run, any particular path looks like a jumble, and the desired
probabilities are not acheived.
```{r, chunk8.1b}
set.seed(100)
KM <- KingMarkov(100)
gf_point(position ~ step, data = KM)
tally( ~ position, data = KM)
```
With a larger number of steps, things are still not perfect, but the are 
much closer to the target probabilities.

```{r, chunk8.1c}
set.seed(123)
KM <- KingMarkov(1e5)
gf_histogram( ~ position, data = KM, binwidth = 1)
KM %>% 
  group_by(position) %>%
  summarise(n = n(), prop = n / 1e5) %>%
  mutate(target = (1:10)/(sum(1:10)))
```

Alternatively, we can propose any island instead of just neighboring islands.

```{r, chunk8.1d}
KingMarkov2 <- function(
  num_steps = 1e5,
  population = 1:10,
  start = 10) {
  
  num_islands <- length(population)
  position <- rep(NA, num_steps)  # trick to pre-alocate memory
  current <- start
  
  for (i in 1:num_steps) {
    # record current position
    position[i] <- current
    
    # propose any one of the other islands
    proposal <- sample(setdiff(1:num_islands, current), 1)
    
    # move?
    prob_move <- population[proposal] / population[current]
    current <- ifelse(runif(1) < prob_move, proposal, current)
  }
  data_frame(
    step = 1:num_steps,
    position = position
  )
}
```

The mixing is faster now.
```{r, chunk8.1e}
set.seed(100)
KM2 <- KingMarkov2(100)
gf_point(position ~ step, data = KM2)
tally( ~ position, data = KM2)
```
With a larger number of steps, things are still not perfect, but the are 
much closer to the target probabilities.

```{r, chunk8.1f}
set.seed(123)
KM2 <- KingMarkov2(1e5)
gf_histogram( ~ position, data = KM2, binwidth = 1)
KM2 %>% 
  group_by(position) %>%
  summarise(n = n(), prop = n / 1e5) %>%
  mutate(target = (1:10)/(sum(1:10)))
```

#### R Code 8.2

```{r, chunk8.2}
library(rethinking)
data(rugged)
Nations <- rugged %>% 
  mutate(log_gdp = log(rgdppc_2000)) %>%
  filter(!is.na(rgdppc_2000))
```

#### R Code 8.3

```{r, chunk8.3}
m8.1 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged + bA * cont_africa + bAR * rugged * cont_africa,
    a ~ dnorm(0, 100),
    bR ~ dnorm(0, 10),
    bA ~ dnorm(0, 10),
    bAR ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = Nations
)
precis(m8.1)
```

#### R Code 8.4

Restrict to 3 variables

```{r, chunk8.4}
Nations.trim <- Nations %>% select(log_gdp, rugged, cont_africa)
str(Nations.trim)
```

#### R Code 8.4

Fit with Stan:
```{r, chunk8.5}
m8.1stan <- map2stan(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged + bA * cont_africa + bAR * rugged * cont_africa,
    a ~ dnorm(0, 100),
    bR ~ dnorm(0, 10),
    bA ~ dnorm(0, 10),
    bAR ~ dnorm(0, 10),
    sigma ~ dcauchy(0, 2)
  ),
  data = Nations.trim
)
```

#### R Code 8.6

```{r, chunk8.6}
precis(m8.1stan)
```

#### R Code 8.7

```{r, chunk8.7}
m8.1stan_4chains <- map2stan(m8.1stan, chains = 4, cores = 4)
precis(m8.1stan_4chains)
```

#### R Code 8.8

```{r, chunk8.8}
m8.1s.post <- extract.samples(m8.1stan)
str(m8.1s.post)
```

#### R Code 8.9

The `rethinking` package includes methods for `pairs()`.

```{r, chunk8.9}
pairs(m8.1s.post)
```

#### R Code 8.10

```{r, chunk8.10}
pairs(m8.1stan)
```

#### R Code 8.11

```{r, chunk8.11}
show(m8.1stan)
```

#### R Code 8.12

```{r, chunk8.12}
plot(m8.1stan)
```

#### R Code 8.13

```{r, chunk8.13}
y <- c(-1, 1)
m8.2 <- map2stan(
  alist(y ~ dnorm(mu, sigma),
        mu <- alpha),
  data = list(y = y),
  start = list(alpha = 0, sigma = 1),
  chains = 2, cores = 2,
  iter = 4000, warmup = 1000
)
```

#### R Code 8.14

```{r, chunk8.14}
precis(m8.2)
plot(m8.2)
stancode(m8.2)
```

#### R Code 8.15

```{r, chunk8.15}
m8.3 <- map2stan(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- alpha,
    alpha ~ dnorm(1, 10),
    sigma ~ dcauchy(0, 1)
  ),
  data = list(y = y),
  start = list(alpha = 0, sigma = 1),
  chains = 2,
  iter = 4000,
  warmup = 1000
)
precis(m8.3)

```

#### R Code 8.16

```{r, chunk8.16}
D <-
  data_frame(
    i = 1:1e4,
    y = rcauchy(1e4, 0, 5),
    mu = cumsum(y) / i
  )
gf_line(mu ~ i, data = D)
```

#### R Code 8.17

```{r, chunk8.17}
y <- rnorm(100, mean = 0, sd = 1)
```

#### R Code 8.18

```{r, chunk8.18}
m8.4 <- map2stan(
  alist(y ~ dnorm(mu, sigma),
        mu <- a1 + a2,
        sigma ~ dcauchy(0, 1)),
  data = list(y = y),
  start = list(a1 = 0, a2 = 0, sigma = 1),
  chains = 2, cores = 2,
  iter = 4000,
  warmup = 1000
)
precis(m8.4)
```

#### R Code 8.19

```{r, chunk8.19}
m8.5 <- map2stan(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- a1 + a2,
    a1 ~ dnorm(0, 10),
    a2 ~ dnorm(0, 10),
    sigma ~ dcauchy(0, 1)
  ),
  data = list(y = y),
  start = list(a1 = 0, a2 = 0, sigma = 1),
  chains = 2, cores = 2,
  iter = 4000, warmup = 1000
)
precis(m8.5)
```

#### R Code 8.20

```{r, chunk8.20}
mp <- map2stan(
  alist(a ~ dnorm(0, 1),
        b ~ dcauchy(0, 1)),
  data = list(y = 1),
  start = list(a = 0, b = 0),
  iter = 1e4,
  warmup = 100,
  WAIC = FALSE
)
```

#### R Code 8.21

Simulating height and leg length.
```{r, chunk8.21}
N <- 100                            # number of individuals
Legs <- 
  data_frame(
    height = rnorm(N, 10, 2),          # sim total height of each
    leg_prop = runif(N, 0.4, 0.5),     # leg as proportion of height
    leg_left  = leg_prop * height + rnorm(N, 0, 0.02),
    leg_right = leg_prop * height + rnorm(N, 0, 0.02)
  )
```

#### R Code 8.22

```{r, chunk8.22}
m5.8s <- map2stan(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * leg_left + br * leg_right,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10),
    sigma ~ dcauchy(0, 1)
  ),
  data = Legs,
  chains = 4, cores = 4,
  start = list(
    a = 10, bl = 0,
    br = 0, sigma = 1
  )
)
```

#### R Code 8

```{r, chunk8.23}
m5.8s2 <- map2stan(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * leg_left + br * leg_right,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10) & T[0, ],
    sigma ~ dcauchy(0, 1)
  ),
  data = Legs,
  chains = 4, cores = 4,
  start = list(
    a = 10, bl = 0,
    br = 0, sigma = 1
  )
)
```
