---
title: "Statistical Rethinking (Code)"
author: "Chapter 12"
date: "April, 2017"
output:
  html_document: 
    fig_height: 3.5
  pdf_document: 
    fig_height: 3.5
---

```{r, setup, include = FALSE}
# Load packages here 
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  cache = TRUE,
  error = TRUE,
  tidy = FALSE,     # display code as typed
  size = "small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_light())

options(`mosaic:parallelMessage` = FALSE)
set.seed(12345)
```


Code from *Statistical Rethinking* modified by R Pruim is shown below.  Differences to the oringal include:

  * a preference for putting data into containers (data frames, mostly), rather than working with lose vectors.
  * use of `ggplot2` (via `ggformula`) rather than base graphics
  * use of `tidyverse` for data transformation
  * better (in my opinion) naming conventions
 
 
### Side note on lists, data frames, and matrices

As our models become more complicated, the containers used to store the data we
use to fit the model and the data we generate from the model (using `link()`,
`sim()`, and `extract.samples()`, for example) also be come more complicated. 
Lists, data frames, and matrices are the containers we will use the most.

  * **lists** are a very general container.  Items in a list are ordered and may be named
  as well.  Each item can contain anything, including data frames, matrices, or other lists.
  Some lists can be converted to data frames using `as.data.frame()`, but this only works
  if the items in the list have the right shape.
  
  * **data frames** are implemented as specialized lists.  This means that you
  can use a data frame in places where are list is required, but not vice versa.
  Each item in a data frame must be of the same length. In most cases each item
  will be a vector. The individual vectors may have different types (character,
  integer, numeric, etc.). It is useful to imagine a data frame as a rectangular
  array with a column for each variable and a row for each case.
 
  * **matrices** are also rectangular, but there is an additional restriction: each item
  must be of the same type (typically numeric or integer).  Rows and columns are numbered
  starting at 1. They may also be named (but often are not).  If `a` is an array, then
  `a[i,j]` is the element in row `i` column `j`; `a[i, ]` is row `i`, and `a[, j]` is 
  column `j`.  `apply()` can be used to apply a function to rows or columns in a matrix.
  The main place we have seen matrices in in the output from `link()` and `sim()`.  
  Since matrices are rectangular, they can be converted to data frames using `as.data.frame()`.
  
  * `str()` can be used to inspect the structure of an object.  This will tell you what
  sort of thing you have and its size.
  
  * **matrix in a list**.  We have seen this before, but it is going to become more common
  with the models in this chapter.  Multi-level models typically include a number of related
  paramters that obtain an index.  In this case `extract.samples()` will return a list of
  posterior samples for each "paramter".  Individual paramters will have a vector 
  of posterior samples.  The related parameters will be stored in a matrix with a column
  for each index and a row for each posterior sample.  `as.data.frame()` will append the index
  onto the base name for these parameters.
  
  * A note about **names**:  It is awkward for R names to include symbols that
  are usually used for syntax -- symbols like `[` and `]` for example.  By
  default, `as.data.frame()` will convert these names, often inserting some dots
  (`.`) in place of the offending symbols.  You can override this default, or
  modify the names later.  To see or modify the names use `names()`, 
  `rownames()`, or  `colnames()`  (If you do want to use names with those symbols, you
  will need to surround them in backquotes each time you use them.)
 
## Reed Frogs

Tadpoles really.  Let's return to our example about survival rates of tadpoles in
various tanks.

#### R Code 12.1

```{r, chunk12.1}
library(rethinking)
data(reedfrogs)
Frogs <- reedfrogs %>% 
  mutate(tank = 1:n())   # make the tank cluster variable
glimpse(Frogs)
```

#### R Code 12.2

```{r, chunk12.2}
# fit
m12.1 <- 
  map2stan(
    alist(
      surv ~ dbinom(density, p),
      logit(p) <- a_tank[tank],
      a_tank[tank] ~ dnorm(0, 5)
    ),
    data = Frogs,
    refresh = 0)
```

#### R Code 12.3

```{r, chunk12.3}
m12.2 <- 
  map2stan(
    alist(
      surv ~ dbinom(density, p),
      logit(p) <- a_tank[tank],
      a_tank[tank] ~ dnorm(a, sigma),
      a ~ dnorm(0, 1),
      sigma ~ dcauchy(0, 1)
    ),
    data = Frogs,
    iter = 4000,
    chains = 4,
    refresh = 0
  )
```

#### R Code 12.4

```{r, chunk12.4}
compare(m12.1, m12.2)
```

#### R Code 12.5

```{r, chunk12.5}
# extract Stan samples
m12.2_post <- extract.samples(m12.2)

# compute median intercept for each tank and transform with logistic()
Frogs <- Frogs %>%
  mutate(
    propsurv.est = logistic(apply(m12.2_post$a_tank, 2, median))
  )
```

```{r, chunk12.5a}
gf_point(propsurv ~ tank, data = Frogs) %>%
  gf_point(propsurv.est ~ tank, data = Frogs, shape = 1) %>%
  # mark posterior median probability across tanks
  gf_hline(yintercept = logistic(median(m12.2_post$a)), lty = 2, alpha = 0.5) %>%
  gf_facet_grid( ~ density, scale="free") %>%
  gf_labs(y = "proportion surviving") 
```

#### R Code 12.6

```{r, chunk12.6}
# show first 100 populations in the posterior
p <-  
  gf_point(y ~ x, data = data_frame(y = 0, x = -4:7), color = "transparent") 

for (i in 1:100) {
  p <- 
    p %>% 
    gf_function(fun = dnorm, 
                args = list(mean = m12.2_post$a[i], sd = m12.2_post$sigma[i]),
                alpha = 0.2)
}
p
```

```{r}
# sample 8000 imaginary tanks from the posterior distribution
sim_tanks <- rnorm(8000, m12.2_post$a, m12.2_post$sigma)

# transform to probability and visualize
gf_dens( ~ (logistic(sim_tanks))) %>%
  gf_labs(x = "probability survive")
```

#### R Code 12.7-12.12



#### R Code 12

```{r, chunk12.9}
FrogSim <-
  function(
    reps = 15,
    n = c(5L, 10L, 25L, 35L),
    a = 1.4,
    sigma = 1.5
    ) {
    nponds <- reps * length(n)
    expand.grid(rep = 1:reps, n = n) %>%
      mutate(
        pond = 1:nponds,
        a_true = rnorm(nponds, mean = a, sd = sigma),
        s = rbinom(nponds, prob = logistic(a_true), size = n)
      )
  }
```

#### R Code 12.10

Stan sometimes requires integer data.  In R, you need to be a bit careful
to ensure that you really have integers and not "numeric" (floating point) values.

```{r, chunk12.10}
class(1:3)
class(c(1, 2, 3))
class(as.integer(c(1, 2, 3)))
class(c(1L, 2L, 3L))  # L for Integer -- go figure
```


#### R Code 12.13

```{r, chunk12.13}
SFrogs <- FrogSim()
m12.3 <- map2stan(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a_pond[pond],
    a_pond[pond] ~ dnorm(a, sigma),
    a ~ dnorm(0, 1),
    sigma ~ dcauchy(0, 1)
  ),
  data = SFrogs,
  iter = 1e4,
  warmup = 1000,
  refresh = 0
)
```

#### R Code 12.14

```{r, chunk12.14}
precis(m12.3, depth = 2)
```

#### R Code 12.15-12.17

By combining this code into a function, we can apply it again later without retyping.  This

  * makes it clearer what is going on, and 
  * makes it easier to make systematic changes.

```{r, chunk12.15}
frog_plot <- function(model, data = model@data) {
  data <-
    data %>% as.data.frame() %>% 
    mutate(
      a_pond_est = as.numeric(coef(model)[1:60]),
      p_est = logistic(a_pond_est),
      p_true = logistic(a_true),
      p_raw = s / n,
      nopool_error = abs(p_raw - p_true),
      partpool_error = abs(p_est - p_true)
    )
  gf_point(nopool_error ~ pond, data = data, shape = 16) %>%
    gf_point(partpool_error ~ pond, data = data, shape = 1) %>%
    gf_line(partpool_error ~ pond, 
            data = data %>% group_by(n) %>%
              mutate(partpool_error = mean(partpool_error)),
            linetype = "dashed", color = "navy") %>%
    gf_line(nopool_error ~ pond, 
            data = data %>% group_by(n) %>%
              mutate(nopool_error = mean(nopool_error)),
            linetype = "dotted", color = "red") %>%
    gf_facet_grid( ~ n, scale = "free") %>%
    gf_labs(y = "absolute error")
}
```

Generating a plot from a model is a one-liner now:

```{r}
frog_plot(model = m12.3)
```

#### R Code 12.19

```{r, chunk12.19}
NewData <- FrogSim()
m12.3new <- 
  map2stan(
    m12.3,
    data = NewData,
    iter = 1e4,
    warmup = 1000)
frog_plot(m12.3new)
```

#### R Code 12

```{r, chunk12.20}
y1 <- rnorm(1e4, 10, 1)
y2 <- 10 + rnorm(1e4, 0, 1)
```

#### R Code 12

```{r, chunk12.21}
library(rethinking)
data(chimpanzees)
Chimps <- chimpanzees %>% select(-recipient)   # get rid of NAs


m12.4 <- map2stan(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <-
      a + a_actor[actor] + (bp + bpC * condition) * prosoc_left,
    a_actor[actor] ~ dnorm(0, sigma_actor),
    a ~ dnorm(0, 10),
    bp ~ dnorm(0, 10),
    bpC ~ dnorm(0, 10),
    sigma_actor ~ dcauchy(0, 1)
  ),
  data = Chimps,
  warmup = 1000,
  iter = 5000,
  chains = 4,
  cores = 3,
  refresh = 0
)
```

#### R Code 12.22

```{r, chunk12.22}
m12.4_post <- extract.samples(m12.4) 
m12.4_post <- 
  m12.4_post %>%
  as.data.frame() %>%
  bind_cols(
    sapply(1:7, function(actor)
      m12.4_post$a + m12.4_post$a_actor[, actor]) %>%
      as.data.frame() %>%
      setNames(paste0("total_a.", 1:7))
  )
```

#### R Code 12.23

```{r, chunk12.23}
# prep data
Chimps <-
  Chimps %>% rename(block_id = block)   # name 'block' is reserved by Stan

m12.5 <- map2stan(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a + a_actor[actor] + a_block[block_id] +
      (bp + bpc * condition) * prosoc_left,
    a_actor[actor] ~ dnorm(0, sigma_actor),
    a_block[block_id] ~ dnorm(0, sigma_block),
    c(a, bp, bpc) ~ dnorm(0, 10),
    sigma_actor ~ dcauchy(0, 1),
    sigma_block ~ dcauchy(0, 1)
  ),
  data = Chimps,
  warmup = 1000,
  iter = 6000,
  chains = 4,
  cores = 3,
  refresh = 0
)
```

#### R Code 12.24

```{r, chunk12.24}
precis(m12.5, depth = 2) # depth=2 displays varying effects
plot(precis(m12.5, depth = 2)) # also plot
```

#### R Code 12.25

```{r, chunk12.25}
m12.5_post <- extract.samples(m12.5) %>% data.frame()
names(m12.5_post)
gf_dens(~sigma_block + color::"block", data = m12.5_post) %>%
  gf_dens(~sigma_actor + color::"actor", data = m12.5_post) %>%
  gf_labs(x = "sigma") %>%
  gf_lims(x = c(0,5))
```

If we convert our posterior samples from a "wide" format (one row per sample) to
a "long" format (one row for each sample of each parameter, we can plot
posterior distributions for multiple parameters even more easily.

```{r, chunk12.25a}
m12.5_post_long <-  m12.5_post %>% 
          tidyr::gather(param, value) 
m12.5_post_long %>% head(3)
gf_dens(~ value + color::param, alpha = 0.6,
        data = m12.5_post_long %>% filter(grepl("sigma", param)))
gf_dens(~ value, 
        data = m12.5_post_long %>% filter(grepl("a_actor\\.", param))) %>%
  gf_vline(xintercept = 0, color = "red", alpha = 0.5, linetype = "dashed") %>%
  gf_facet_grid( param ~ .)
gf_dens(~ value,
        data = m12.5_post_long %>% filter(grepl("a_block\\.", param))) %>%
  gf_vline(xintercept = 0, color = "red", alpha = 0.5, linetype = "dashed") %>%
  gf_facet_grid( param ~ .)
```

#### R Code 12.26

```{r, chunk12.26}
compare(m12.4, m12.5)
```

#### R Code 12.27

```{r, chunk12.27}
m12.4_pred <-
  expand.grid(
    prosoc_left = 0:1,
    condition = 0:1,
    actor = 1:7) %>%
  mutate(
    combo = paste0(prosoc_left, "/", condition)
  )
  
link.m12.4 <- link(m12.4, data = m12.4_pred)
m12.4_pred <- m12.4_pred %>% 
  mutate(
    p.pred = apply(link.m12.4, 2, mean),
    p.link.lo = apply(link.m12.4, 2, PI)[1,],
    p.link.hi = apply(link.m12.4, 2, PI)[2,]
  )
```

```{r}
gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = m12.4_pred) %>%
  gf_line(p.pred ~ combo + group::"1", data = m12.4_pred) %>%
  gf_facet_wrap( ~ actor)
```

#### R Code 12.28

```{r, chunk12.28}
m12.4_post <- extract.samples(m12.4) 
m12.4_postD <- m12.4_post %>% data.frame()
str(m12.4_post)
str(m12.4_postD)
```

#### R Code 12.29

```{r, chunk12.29}
gf_dens( ~ a_actor.5, data = m12.4_postD)
m12.4_post_long <- 
  m12.4_postD %>%
  tidyr::gather(param, value)
gf_dens( ~ value, data = m12.4_post_long %>% filter(grepl("actor\\.", param))) %>%
  gf_vline(xintercept = 0, color = "red", alpha = 0.5, linetype = "dashed") %>%
  gf_facet_grid( param ~ .)
```

### Creating the link function manually

For some models, it is necessary to create posterior values manually rather than with 
`link()` or `sim()`.  Although this is not required here, we illustrate the manual approach
by way of comparison.

#### R Code 12.30

```{r, chunk12.30}
p.link <- function(prosoc_left, condition, actor, post = m12.4_post) {
  logodds <- 
    with(post,
         a + a_actor[, actor] + (bp + bpC * condition) * prosoc_left)
  return(logistic(logodds))
} 
str(p.link(0, 0, 2))
```

#### R Code 12.31

```{r, chunk12.31}
G <- expand.grid(
  prosoc_left = 0:1,
  condition = 0:1,
  actor = 1:7) %>%
  mutate(
    combo = paste0(prosoc_left, "/", condition)
  )

manual_link <-
  with(G,    # saves having to type G$ repeatedly
       mapply(p.link, prosoc_left, condition, actor)
  ) 
# Note: a column here for each row in G
str(manual_link)
dim(G)

manual12.4_pred <- 
  G %>%
  mutate(
    p.pred = apply(manual_link, 2, mean),
    p.link.lo = apply(manual_link, 2, PI)[1,],
    p.link.hi = apply(manual_link, 2, PI)[2,]
  )
```

```{r}
# replication of plot from above using home spun link
gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = manual12.4_pred) %>%
  gf_line(p.pred ~ combo + group::"1", data = m12.4_pred) %>%
  gf_facet_wrap( ~ actor)
```


#### R Code 12.32

```{r, chunk12.32}
# don't need multiple actors this time
G <- expand.grid(
  prosoc_left = 0:1,
  condition = 0:1,
  actor = 1) %>%
  mutate(
    combo = paste0(prosoc_left, "/", condition)
  )
```

#### R Code 12.33

```{r, chunk12.33}
# replace varying intercept samples with zeros
# 1000 samples by 7 actors
a_actor_zeros <- matrix(0, 1000, 7)
```

#### R Code 12.34

```{r, chunk12.34}
# note use of replace list
m12.4_link2 <- 
  link(
    m12.4,
    n = 1000,
    data = G,
    replace = list(a_actor = a_actor_zeros)
  )

m12.4_pred2 <- 
  G %>% 
  mutate(
    p.pred = apply(m12.4_link2, 2, mean),
    p.link.lo = apply(m12.4_link2, 2, PI, prob = 0.8)[1,],
    p.link.hi = apply(m12.4_link2, 2, PI, prob = 0.8)[2,]
  )


gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = m12.4_pred2) %>%
  gf_line(p.pred ~ combo + group::"doesn't matter what this is", data = m12.4_pred2) %>%
  gf_labs(x = "prosoc_left/condition", y = "proportion pulled left", title = "average actor")
```

#### R Code 12.35

```{r, chunk12.35}
# replace varying intercept samples with simulations
m12.4_post <- extract.samples(m12.4)
a_actor_sims <- 
  rnorm(7000, 0, m12.4_post$sigma_actor) %>%
  matrix(1000, 7)   # reshape into a 1000 x 7 matrix
```

```{r}
m12.4_pred3 <- 
  G %>% 
  mutate(
    p.pred = apply(m12.4_link3, 2, mean),
    p.link.lo = apply(m12.4_link3, 2, PI, prob = 0.8)[1,],
    p.link.hi = apply(m12.4_link3, 2, PI, prob = 0.8)[2,]
  )


gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = m12.4_pred3) %>%
  gf_line(p.pred ~ combo + group::"doesn't matter what this is", data = m12.4_pred3) %>%
  gf_labs(x = "prosoc_left/condition", y = "proportion pulled left", title = "marginal of actor")
```

#### R Code 12.36

```{r, chunk12.36}
m12.4_link3 <- 
  link(
    m12.4,
    n = 1000,
    data = G,
    replace = list(a_actor = a_actor_sims)
  )
```

#### R Code 12.37

```{r, chunk12.37}
m12.4_post <- extract.samples(m12.4) %>%
  as.data.frame() %>%
  mutate(
    sim_a_actor = rnorm(16000, 0, sigma_actor)
  )

Actors50 <- 
  expand.grid(
    actor = 1:50,        # 50 simulated actors
    prosoc_left = 0:1,
    condition = 0:1) %>% 
  mutate(
    combo = paste0(prosoc_left, "/", condition),
    logodds = 
      m12.4_post$a[actor] +
      m12.4_post$sim_a_actor[actor] +
      (m12.4_post$bp[actor] + m12.4_post$bpC[actor] * condition) * prosoc_left,
    p = logistic(logodds)
  )
```

#### R Code 12.38

```{r, chunk12.38}
gf_line(p ~ combo + group::actor, data = Actors50, alpha = 0.3) %>%
  gf_labs(
    x = "prosoc_left/condition",
    y = "proportion pulled left",
    title = "50 simulated actors"
  )
```


## Kline data

#### R Code 12.39

```{r, chunk12.39}
# prep data
library(rethinking)
data(Kline)
Kline  <- Kline %>%
  mutate(
    logpop = log(population),
    society = 1:n()
  )
```

```{r}
# fit model
m12.6 <- map2stan(
  alist(
    total_tools ~ dpois(mu),
    log(mu) <- a + a_society[society] + bp * logpop,
    a ~ dnorm(0, 10),
    bp ~ dnorm(0, 1),
    a_society[society] ~ dnorm(0, sigma_society),
    sigma_society ~ dcauchy(0, 1)
  ),
  data = Kline,
  iter = 4000,
  chains = 3,
  refresh = 0
)
```

#### R Code 12.40

```{r, chunk12.40}
m12.6_post <- extract.samples(m12.6)
m12.6_pred <- 
  expand.grid(
    logpop = seq(from = 6, to = 14, length.out = 30),
    society = 1)
a_society_sims <- 
  rnorm(20000, 0, m12.6_post$sigma_society) %>%
  matrix(2000, 10)

m12.6_link <- 
  link(
    m12.6,
    n = 2000,
    data = m12.6_pred,
    replace = list(a_society = a_society_sims)
  )
```

#### R Code 12.41

```{r, chunk12.41}
m12.6_pred <- 
  m12.6_pred %>%
  mutate(
    mu.median = apply(m12.6_link, 2, median),
    link67.lo = apply(m12.6_link, 2, PI, prob = 0.67)[1,],
    link67.hi = apply(m12.6_link, 2, PI, prob = 0.67)[2,],
    link89.lo = apply(m12.6_link, 2, PI, prob = 0.89)[1,],
    link89.hi = apply(m12.6_link, 2, PI, prob = 0.89)[2,],
    link97.lo = apply(m12.6_link, 2, PI, prob = 0.97)[1,],
    link97.hi = apply(m12.6_link, 2, PI, prob = 0.97)[2,]
  )
    
gf_ribbon(link67.lo + link67.hi ~ logpop, data = m12.6_pred, alpha = 0.1) %>%
gf_ribbon(link89.lo + link89.hi ~ logpop, data = m12.6_pred, alpha = 0.1) %>%
gf_ribbon(link97.lo + link97.hi ~ logpop, data = m12.6_pred, alpha = 0.1) %>%
gf_point(total_tools ~ logpop, data = Kline) %>%
  gf_labs(x = "log population", y = "total tools") %>%
  gf_line(mu.median ~ logpop, data = m12.6_pred)
```

#### R Code 12.42

There is one missing district number
```{r, chunk12.42}
data(bangladesh)
# would you have noticed that 54 is missing in this list?
sort(unique(bangladesh$district))
# it is easier to spot the 2 in this list
diff(sort(unique(bangladesh$district)))
# or we can use a plot to spot that there is one district number missing
gf_histogram(~ district, data = Bangladesh, binwidth = 1)
```

#### R Code 12.43

Here are two ways to convert to a proper index.  Notice that the results are not the same.
The first is perhaps preferable since it retains the order of the original district numbers.
(In the second one, 2, 3, 4, etc. come just before 20, 30, 40, etc. because "alphabetical" order
is used rather than numerical order.

```{r, chunk12.43}
Bangladesh <-
  bangladesh %>% 
  mutate(
    district_id1 = as.integer(factor(district)),
    district_id2 = coerce_index(district)
  )
diff(sort(unique(Bangladesh$district_id1)))
diff(sort(unique(Bangladesh$district_id2)))
tally(~ (district_id1 == district_id2), data = Bangladesh)
gf_abline(intercept = 0, slope = 1, alpha = 0.4) %>%
gf_point(district_id2 ~ district + color::"2 vs 0", data = Bangladesh) %>%
  gf_point(district_id1 ~ district + color::"1 vs 0", data = Bangladesh) %>%
  gf_labs(title = "Comparing two conversions to an index")
```
