---
title: "Statistical Rethinking (Code)"
author: "Chapter 4"
date: "February, 2017"
output:
  html_document: default
  pdf_document: default
params:
  original: no
  updated: yes
---

```{r, opts.label = "original", results = "asis", echo = FALSE}
cat("The original code from *Statistical Rethinking* is shown below.")
cat("\n\n")
```

```{r, opts.label = "updated", results = "asis", echo = FALSE}
cat(paste(
"Code from *Statistical Rethinking* modified by R Pruim is shown below.  Differences to the oringal include:",
"",
"  * a preference for putting data into containers (data frames, mostly), rather than working with lose vectors.",
"  * use of `lattice` and `ggplot2` rather than base graphics",
"  * use of `tidyverse` for data transformation",
"  * better (in my opinion) naming conventions",
sep = "\n"
))
cat("\n\n")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(mosaic)
require(rethinking)

knitr::opts_template$set(
  original = list(eval = params$original, include = params$original),
  updated = list(eval = params$updated, include = params$updated))

trellis.par.set(theme = theme.mosaic())
theme_set(theme_minimal())
set.seed(123)   # get same random draws each time the file is processed.
```


## R code 4.1

```{r, chunk4.1}
pos <- replicate(1000, sum(runif(16, -1, 1)))
```

## R code 4.2

```{r, chunk4.2}
prod(1 + runif(12, 0, 0.1))
```

## R code 4.3

```{r, chunk4.3, opts.label = "original"}
growth <- replicate(10000, prod(1 + runif(12, 0, 0.1)))
dens(growth, norm.comp = TRUE)
```

```{r, chunk4.3a, opts.label = "updated"}
growth <- replicate(10000, prod(1 + runif(12, 0, 0.1)))
densityplot(~ growth)
```

## R code 4.4

```{r, chunk4.4}
big <- replicate(10000, prod(1 + runif(12, 0, 0.5)))
small <- replicate(10000, prod(1 + runif(12, 0, 0.01)))
```

## R code 4.5

```{r, chunk4.5}
log.big <- replicate(10000, log(prod(1 + runif(12, 0, 0.5))))
```

## R code 4.6

```{r, chunk4.6, opts.label = "original"}
w <- 6
n <- 9

p_grid <- seq(from = 0,
              to = 1,
              length.out = 100)
posterior <- dbinom(w, n, p_grid) * dunif(p_grid, 0, 1)
posterior <- posterior / sum(posterior)
```

```{r, chunk4.6a, opts.label = "updated"}
w <- 6
n <- 9

Water9 <-
  expand.grid(
    p = seq(from = 0, to = 1, by = 0.01)) %>%
  mutate(prior = dunif(p, 0, 1),
         likelihood = dbinom(w, n, p),
         posterior.raw = prior * likelihood,
         posterior1 = posterior.raw / sum(posterior.raw),
         posterior = posterior1 / 0.01
  )

xyplot(prior + posterior ~ p, data = Water9, type = "l")
```

## R code 4.7

```{r, chunk4.7, opts.label = "original"}
library(rethinking)
data(Howell1)
d <- Howell1
```

```{r, chunk4.7a, opts.label = "updated"}
library(rethinking)
data(Howell1)
```


## R code 4.8

```{r, chunk4.8, opts.label = "original"}
str(d)
```

```{r, chunk4.8a, opts.label = "updated"}
str(Howell1)
glimpse(Howell1)
```

## R code 4.9

```{r, chunk4.9, results = "hide", opts.label = "original"}
d$height
```

```{r, chunk4.9a, results = "hide"}
Howell1$height
```

## R code 4.10

```{r, chunk4.10, opts.label = "original"}
d2 <- d[d$age >= 18,]
```

```{r, chunk4.10a, opts.label = "updated"}
HowellAdults <- Howell1 %>% filter(age >= 18)
```


## R code 4.11

```{r, chunk4.11, opts.label = "original"}
curve(dnorm(x, 178, 20), from = 100, to = 250)
```

```{r, chunk4.11a, opts.label = "updated"}
plotDist("norm", mean = 178, sd = 20)
```


## R code 4.12

```{r, chunk4.12, opts.label = "original"}
curve(dunif(x, 0, 50), from = -10, to = 60)
```

```{r, chunk4.12a, opts.label = "updated"}
plotDist("unif", min = 0, max = 50)
```

## R code 4.13

```{r, chunk4.13, opts.label = "original"}
sample_mu <- rnorm(1e4, 178, 20)
sample_sigma <- runif(1e4, 0, 50)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)
```

```{r, chunk4.13a, opts.label = "updated"}
PriorSample <- data_frame(
  mu     = rnorm(1e4, 178, 20),
  sigma  = runif(1e4, 0, 50),
  height = rnorm(1e4, mu, sigma)
)
densityplot( ~ height, data = PriorSample)
```

## R code 4.14

```{r, chunk4.14, opts.label = "original"}
mu.list <- seq(from = 140,
               to = 160,
               length.out = 200)   # Note: this is a vector in R
sigma.list <- seq(from = 4,
                  to = 9,
                  length.out = 200)  # This is also a vector

post <- expand.grid(mu = mu.list, sigma = sigma.list)
post$LL <- sapply(1:nrow(post), function(i)
  sum(dnorm(
    d2$height,
    mean = post$mu[i],
    sd = post$sigma[i],
    log = TRUE
  )))
post$prod <- post$LL + dnorm(post$mu, 178, 20, TRUE) +
  dunif(post$sigma, 0, 50, TRUE)
post$prob <- exp(post$prod - max(post$prod))
```



```{r, chunk4.14a, opts.label = "updated"}
llnorm <- function(x, mu, sigma) { 
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}
# use all of x for each mu-sigma combo
llnorm <- Vectorize(llnorm, c("mu", "sigma"))
                
HowellAdultsGrid <-
  expand.grid(
    mu = seq(from = 140, to = 160, length.out = 200), 
    sigma = seq(from = 4, to = 9, length.out = 200)) %>% 
  mutate(
    logprior = 
      dnorm(mu, 178, 20, log = TRUE) +
      dunif(sigma, 0, 50, log = TRUE),
    loglik = llnorm(HowellAdults$height, mu, sigma),
    posterior.log = logprior + loglik,
    posterior.raw = exp(posterior.log - max(posterior.log)),
    posterior = posterior.raw / sum(posterior.raw)
  )
```


## R code 4.15

```{r, chunk4.15, opts.label = "original"}
contour_xyz(post$mu, post$sigma, post$prob)
```

```{r, chunk4.15a, opts.label = "updated"}
contourplot(posterior ~ mu + sigma, data = HowellAdultsGrid)

```

## R code 4.16

```{r, chunk4.16, opts.label = "original"}
image_xyz(post$mu, post$sigma, post$prob)
```

```{r, chunk4.16a, opts.label = "updated"}
levelplot(posterior ~ mu + sigma, data = HowellAdultsGrid, contour = TRUE)
levelplot(posterior.log ~ mu + sigma, data = HowellAdultsGrid, contour = TRUE)
```
## R code 4.17

```{r, chunk4.17, opts.label = "original"}
sample.rows <- sample(1:nrow(post),
                      size = 1e4,
                      replace = TRUE,
                      prob = post$prob)
sample.mu <- post$mu[sample.rows]
sample.sigma <- post$sigma[sample.rows]
```

```{r, chunk4.17a, opts.label = "updated"}
PosteriorSample <- 
  sample(HowellAdultsGrid, size = 1e4, replace = TRUE, 
         prob = HowellAdultsGrid$posterior)
```

## R code 4.18

```{r, chunk4.18, opts.label = "original"}
plot(
  sample.mu,
  sample.sigma,
  cex = 0.5,
  pch = 16,
  col = col.alpha(rangi2, 0.1)
)
```

```{r, chunk4.18a, opts.label = "updated"}
xyplot(sigma ~  mu, data = PosteriorSample,
  cex = 0.5,
  pch = 16,
  col = col.alpha(rangi2, 0.1)
)

ggplot(PosteriorSample, aes(x = mu, y = sigma)) +
  geom_point(alpha = 0.2) +
  geom_density2d() 
       
ggplot(PosteriorSample, aes(x = mu, y = sigma)) +
  geom_hex() 
       
```

## R code 4.19

```{r, chunk4.19, opts.label = "original"}
dens(sample.mu)
dens(sample.sigma)
```

```{r, chunk4.19a, opts.label = "updated"}
densityplot( ~ mu, data = PosteriorSample)
densityplot( ~ sigma, data = PosteriorSample)
```

## R code 4.20

```{r, chunk4.20, opts.label = "original"}
HPDI(sample.mu)
HPDI(sample.sigma)
```

```{r, chunk4.20a, opts.label = "updated"}
HPDI(PosteriorSample$mu)
HPDI(PosteriorSample$sigma)
```

## R code 4.21

```{r, chunk4.21, opts.label = "original"}
d3 <- sample(d2$height, size = 20)
```

```{r, chunk4.21a, opts.label = "updated"}
# sample only heights
sample(HowellAdults$height, size = 20)
# sample complete rows
sample(HowellAdults, size = 20)
```

## R code 4.22

```{r, chunk4.22, opts.label = "original"}
mu.list <- seq(from = 150,
               to = 170,
               length.out = 200)
sigma.list <- seq(from = 4,
                  to = 20,
                  length.out = 200)
post2 <- expand.grid(mu = mu.list, sigma = sigma.list)
post2$LL <- sapply(1:nrow(post2), function(i)
  sum(dnorm(
    d3,
    mean = post2$mu[i],
    sd = post2$sigma[i],
    log = TRUE
  )))
post2$prod <- post2$LL + dnorm(post2$mu, 178, 20, TRUE) +
  dunif(post2$sigma, 0, 50, TRUE)
post2$prob <- exp(post2$prod - max(post2$prod))
sample2.rows <- sample(
  1:nrow(post2),
  size = 1e4,
  replace = TRUE,
  prob = post2$prob
)
sample2.mu <- post2$mu[sample2.rows]
sample2.sigma <- post2$sigma[sample2.rows]
plot(
  sample2.mu,
  sample2.sigma,
  cex = 0.5,
  col = col.alpha(rangi2, 0.1),
  xlab = "mu",
  ylab = "sigma",
  pch = 16
)
```

```{r, chunk4.22a, opts.label = "updated"}
llnorm <- function(x, mu, sigma) { 
  sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
}
# use all of x for each mu-sigma combo
llnorm <- Vectorize(llnorm, c("mu", "sigma"))

Howell20 <- sample(HowellAdults, 20)
                
Howell20Grid <-
  expand.grid(
    mu = seq(from = 140, to = 160, by = 0.1),
    sigma = seq(from = 2, to = 12, by = 0.1)) %>% 
  mutate(
    logprior = 
      dnorm(mu, 178, 20, log = TRUE) +
      dunif(sigma, 0, 50, log = TRUE),
    loglik = llnorm(Howell20$height, mu, sigma),
    posterior.log = logprior + loglik,
    posterior.raw = exp(posterior.log - max(posterior.log)),
    posterior = posterior.raw / sum(posterior.raw)
  )

Howell20Posterior <- 
  sample(Howell20Grid, size = 1e4,
         replace = TRUE, 
         prob = Howell20Grid$posterior)

xyplot(sigma ~ mu, data = Howell20Posterior,
  cex = 0.5, alpha = 0.1,
  xlab = expression(mu), ylab = expression(sigma),
  pch = 16
)

ggplot(Howell20Posterior, aes(x = mu, y = sigma)) +
  geom_point(alpha = 0.1) + 
  geom_density2d() +
  labs(x = expression(mu), y = expression(sigma))
```

## R code 4.23

```{r, chunk4.23, opts.label = "original"}
dens(sample2.sigma, norm.comp = TRUE)
```

```{r, chunk4.23a, opts.label = "updated"}
densityplot( ~ sigma, data = Howell20Posterior)
histogram( ~ sigma, data = Howell20Posterior, width = 0.1, fit = "normal")
xqqmath( ~ sigma, data = Howell20Posterior)
```

## R code 4.24

```{r, chunk4.24, opts.label = "original"}
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[d$age >= 18, ]
```

```{r, chunk4.24a, opts.label = "updated"}
library(rethinking)
data(Howell1)
HowellAdults <- Howell1 %>% filter(age >= 18)
```

## R code 4.25

```{r, chunk4.25}
flist <- 
  alist(
    height ~ dnorm(mu, sigma),
    mu ~ dnorm(178, 20),
    sigma ~ dunif(0, 50))
```

## R code 4.26

```{r, chunk4.26, opts.label = "original"}
m4.1 <- map(flist, data = d2)
```

```{r, chunk4.26a, opts.label = "updated"}
m4.1 <- map(flist, data = HowellAdults)
```

## R code 4.27

```{r, chunk4.27}
precis(m4.1)
```

## R code 4.28

```{r, chunk4.28, opts.label = "original"}
start <- list(mu = mean(d2$height),
              sigma = sd(d2$height))
```

```{r, chunk4.28a, opts.label = "updated"}
start <- 
  list(
    mu  = mean( ~ height, data = HowellAdults),
    sigma = sd( ~ height, data = HowellAdults))
```

## R code 4.29

```{r, chunk4.29, opts.label = "original"}
m4.2 <- map(alist(height ~ dnorm(mu, sigma),
                      mu ~ dnorm(178, 0.1),
                   sigma ~ dunif(0, 50)),
            data = d2)
precis(m4.2)
```

```{r, chunk4.29a, opts.label = "updated"}
m4.2 <- map(alist(height ~ dnorm(mu, sigma),
                      mu ~ dnorm(178, 0.1),
                   sigma ~ dunif(0, 50)),
            data = HowellAdults)
precis(m4.2)
```

## R code 4.30

```{r, chunk4.30}
vcov(m4.1)
```

## R code 4.31

```{r, chunk4.31}
diag(vcov(m4.1))
cov2cor(vcov(m4.1))
```

## R code 4.32

```{r, chunk4.32}
library(rethinking)
post <- extract.samples(m4.1, n = 1e4)
head(post)
```

## R code 4.33

```{r, chunk4.33}
precis(post)
```

## R code 4.34

```{r, chunk4.34}
library(MASS)
post <- mvrnorm(n = 1e4,
                mu = coef(m4.1),
                Sigma = vcov(m4.1))
```

## R code 4.5

```{r, chunk4.35, opts.label = "original"}
m4.1_logsigma <- map(alist(
  height ~ dnorm(mu, exp(log_sigma)),
  mu ~ dnorm(178, 20),
  log_sigma ~ dnorm(2, 10)
), data = d2)
```

```{r, chunk4.35a, opts.label = "updated"}
m4.1_logsigma <- 
  map(
    alist(
      height ~ dnorm(mu, exp(log_sigma)),
      mu ~ dnorm(178, 20),
      log_sigma ~ dnorm(2, 10)
    ), data = HowellAdults)
```

## R code 4.36

```{r, chunk4.36, opts.label = "original"}
post <- extract.samples(m4.1_logsigma)
sigma <- exp(post$log_sigma)
```


```{r, chunk4.36a, opts.label = "updated"}
m4.1_logsimga.post <- 
  extract.samples(m4.1_logsigma) %>%
  mutate(sigma = exp(log_sigma))
```

# Section 4.4

## R code 4.37

```{r, chunk4.37, opts.label = "original"}
plot(d2$height ~ d2$weight)
```

```{r, chunk4.37a, opts.label = "updated"}
xyplot(height ~ weight, data = HowellAdults)
```

## R code 4.38

```{r, chunk4.38, opts.label = "original"}
# load data again, since it's a long way back
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[d$age >= 18,]

# fit model
m4.3 <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- a + b * weight,
  a ~ dnorm(156, 100),
  b ~ dnorm(0, 10),
  sigma ~ dunif(0, 50)
),
data = d2)
```


```{r, chunk4.38a, opts.label = "updated"}
# load data again, since it's a long way back
library(rethinking)
data(Howell1)
HowellAdults <-
  Howell1 %>% filter(age > 18)

# fit model
m4.3 <- 
  map(
    alist(
      height ~ dnorm(mu, sigma),
      mu <- a + b * weight,
      a ~ dnorm(156, 100),
      b ~ dnorm(0, 10),
      sigma ~ dunif(0, 50)
    ),
    data = HowellAdults)
```

## R code 4.39

```{r, chunk4.39, opts.label = "original"}
# but note: this doesn't work with link()!
m4.3a <- map(alist(
  height ~ dnorm(a + b * weight, sigma),
  a ~ dnorm(178, 100),
  b ~ dnorm(0, 10),
  sigma ~ dunif(0, 50)
),
data = d2)
```

```{r, chunk4.39a, opts.label = "updated"}
# but note: this doesn't work with link()!
m4.3a <- map(alist(
  height ~ dnorm(a + b * weight, sigma),
  a ~ dnorm(178, 100),
  b ~ dnorm(0, 10),
  sigma ~ dunif(0, 50)
),
data = HowellAdults)
```

## R code 4.40

```{r, chunk4.40}
precis(m4.3)
```

## R code 4.41

```{r, chunk4.41}
precis(m4.3, corr = TRUE)
```

## R code 4.42

Centering `weight`:

```{r, chunk4.42, opts.label = "original"}
d2$weight.c <- d2$weight - mean(d2$weight)
```

```{r, chunk4.42a, opts.label = "updated"}
HowellAdults <- 
  HowellAdults %>% 
  mutate(weight.c = weight - mean(weight))
```

## R code 4.43

```{r, chunk4.43, opts.label = "original"}
m4.4 <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- a + b * weight.c,
  a ~ dnorm(178, 100),
  b ~ dnorm(0, 10),
  sigma ~ dunif(0, 50)
),
data = d2)
```

```{r, chunk4.43a, opts.label = "updated"}
m4.4 <- 
  map(
    alist(
      height ~ dnorm(mu, sigma),
      mu <- a + b * weight.c,
      a ~ dnorm(178, 100),
      b ~ dnorm(0, 10),
      sigma ~ dunif(0, 50)
    ),
    data = HowellAdults)
```

## R code 4.44

```{r, chunk4.44}
precis(m4.4, corr = TRUE)
```

## R code 4.45

```{r, chunk4.45, opts.label = "original"}
plot(height ~ weight, data = d2)
abline(a = coef(m4.3)["a"], b = coef(m4.3)["b"])
```

```{r, chunk4.45a, opts.label = "updated"}
# lattice with custom panel function to overlay multiple things
xyplot(height ~ weight, data = HowellAdults,
       panel = function(x, y, ...) {
         panel.abline(coef(m4.3))
         panel.xyplot(x, y, ...)
       }
)
```

```{r, chunk4.45b, opts.label = "updated", fig.keep = "last"}
# using plotFun() with add = TRUE to plot a function on top of points
xyplot(height ~ weight, data = HowellAdults)
plotFun(a + b * x ~ x, a = coef(m4.3)["a"], b = coef(m4.3)["b"], add = TRUE, col = "red")
```

```{r, chunk4.45c, opts.label = "updated"}
# creating a separate function for use with plotFun()
xyplot(height ~ weight, data = HowellAdults)
line.fit <- function(x, a = coef(m4.3)["a"], b = coef(m4.3)["b"]) {
  a + b * x
}
plotFun(line.fit(height) ~ height, add = TRUE, col = "red")
```

## R code 4.46

```{r, chunk4.46, opts.label = "original"}
post <- extract.samples(m4.3)
```

```{r, chunk4.46a, opts.label = "updated"}
m4.3.post <- extract.samples(m4.3)
```

## R code 4.47

```{r, chunk4.47, opts.label = "original"}
post[1:5, ]
```

```{r, chunk4.47a, opts.label = "updated"}
head(m4.3.post)
```
 ## R code 4.48

```{r, chunk4.48, opts.label = "original"}
N <- 10
dN <- d2[1:N,]
mN <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- a + b * weight,
  a ~ dnorm(178, 100),
  b ~ dnorm(0, 10),
  sigma ~ dunif(0, 50)
),
data = dN)
```

```{r, chunk4.48a, opts.label = "updated"}
HowellSmall <- Howell1 %>% sample(10)
m4.4small <- 
  map(alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b * weight,
    a ~ dnorm(178, 100),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = HowellSmall)
```

## R code 4.49

```{r, chunk4.49, opts.label = "original"}
# extract 20 samples from the posterior
post <- extract.samples(mN, n = 20)

# display raw data and sample size
plot(
  dN$weight,
  dN$height,
  xlim = range(d2$weight),
  ylim = range(d2$height),
  col = rangi2,
  xlab = "weight",
  ylab = "height"
)
mtext(concat("N = ", N))

# plot the lines, with transparency
for (i in 1:20)
  abline(a = post$a[i],
         b = post$b[i],
         col = col.alpha("black", 0.3))

```

```{r, chunk4.49a, opts.label = "updated"}
# extract 50 samples from the posterior
m4.4small.post <- extract.samples(m4.4small, n = 50)

# display raw data and sample size
xyplot(height ~ weight, data = HowellSmall,
       xlim = c(30, 65),
       ylim = c(130, 180),
       col = rangi2,
       main = concat("N = ", 10),
       panel = function(x, y, ...) {
         panel.xyplot(x, y, ...)
         # plot the lines, with transparency
         for (i in 1:nrow(m4.4small.post)) {
           panel.abline(a = m4.4small.post$a[i], b = m4.4small.post$b[i], col = "black",
                        alpha = 0.1)
         }
       }
)
```


## R code 4.50

```{r, chunk4.50, opts.label = "original"}
mu_at_50 <- post$a + post$b * 50
```

```{r, chunk4.50a, opts.label = "updated"}
m4.4small.post <-
  m4.4small.post %>%
  mutate(mu_at_50 = a + b * 50)
```

## R code 4.51

```{r, chunk4.51, opts.label = "original"}
dens(mu_at_50,
     col = rangi2,
     lwd = 2,
     xlab = "mu|weight=50")
```

```{r, chunk4.51a, opts.label = "updated"}
densityplot( ~ mu_at_50, data = m4.4small.post,
  col = rangi2, lwd = 2, xlab = expression(paste(mu, " | weight=50")))
```

## R code 4.52

```{r, chunk4.52, opts.label = "original"}
HPDI(mu_at_50, prob = 0.89)
```

```{r, chunk4.52a, opts.label = "updated"}
HPDI(m4.4small.post$mu_at_50, prob = 0.89)
```


## R code 4.53

```{r, chunk4.53}
mu <- link(m4.3)
str(mu)
```

## R code 4.54

```{r, chunk4.54, opts.label = "original"}
# define sequence of weights to compute predictions for
# these values will be on the horizontal axis
weight.seq <- seq(from = 25, to = 70, by = 1)

# use link to compute mu
# for each sample from posterior
# and for each weight in weight.seq
mu <- link(m4.3, data = data.frame(weight = weight.seq))
str(mu)
```


```{r, chunk4.54a, opts.label = "updated"}
# define sequence of weights to compute predictions for
# these values will be on the horizontal axis
m4.3.pred <-
  data.frame(weight = seq(from = 25, to = 70, by = 1))

# use link to compute mu
# for each sample from posterior
# and for each weight in weight.seq
mu <- 
  link(
    m4.3, 
    data = m4.3.pred,
  )
str(mu)
```

## R code 4.55

```{r, chunk4.55, opts.label = "original"}
# use type="n" to hide raw data
plot(height ~ weight, d2, type = "n")

# loop over samples and plot each mu value
for (i in 1:100)
  points(weight.seq, mu[i, ], pch = 16, col = col.alpha(rangi2, 0.1))
```

```{r, chunk4.55a, fig.keep = "last"}
# The shape of mu coming out of link() is backwards for lattice
# t() flips rows and columns to make it work
xyplot(t(mu) ~ weight, data = m4.3.pred, alpha = 0.1, col = rangi2, pch = 16)
```

## R code 4.56

```{r, chunk4.56}
# summarize the distribution of mu
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.89)
```

## R code 4.57

```{r, chunk4.57, opts.label = "original"}
# plot raw data
# fading out points to make line and interval more visible
plot(height ~ weight, data = d2, col = col.alpha(rangi2, 0.5))

# plot the MAP line, aka the mean mu for each weight
lines(weight.seq, mu.mean)

# plot a shaded region for 89% HPDI
shade(mu.HPDI, weight.seq)
```


```{r, chunk4.57a, fig.keep = "last"}
# plot raw data
# fading out points to make line and interval more visible
xyplot(height ~ weight, data = HowellAdults, col = rangi2, alpha = 0.5)

# plot the MAP line, aka the mean mu for each weight
plotPoints(mu.mean ~ weight, data = m4.3.pred, type = "l", add = TRUE)

# plot a shaded region for 89% HPDI
# need to figure out easiest way to do this in lattice
# author's shade() function only works with his base graphics
```


## R code 4.58

Here's roughly how `link()` works:

```{r, chunk4.58}
post <- extract.samples(m4.3)
mu.link <- function(weight)
  post$a + post$b * weight
weight.seq <- seq(from = 25, to = 70, by = 1)
mu <- sapply(weight.seq, mu.link)
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.89)
```

## R code 4.59

```{r, chunk4.59, opts.label = "original"}
sim.height <- sim(m4.3, data = list(weight = weight.seq))
str(sim.height)
```

```{r, chunk4.59a, opts.label = "updated"}

sim.height <- sim(m4.3, data = m4.3.pred)  
str(sim.height)
```

## R code 4.60

```{r, chunk4.60}
height.PI <- apply(sim.height, 2, PI, prob = 0.89)
```

## R code 4.61

```{r, chunk4.61, opts.label = "original"}
# plot raw data
plot(height ~ weight, d2, col = col.alpha(rangi2, 0.5))

# draw MAP line
lines(weight.seq, mu.mean)

# draw HPDI region for line
shade(mu.HPDI, weight.seq)

# draw PI region for simulated heights
shade(height.PI, weight.seq)
```

```{r, chunk4.61a, fig.keep = "last"}
# plot raw data
xyplot(height ~ weight, HowellAdults, col = rangi2, alpha = 0.5)

# draw MAP line
plotPoints(mu.mean ~ weight, data = m4.3.pred, 
           col = "blue", type = "l", add = TRUE)

# need a replacement for shade() that works with lattice.
# draw HPDI region for line
# shade(mu.HPDI, weight.seq)
plotPoints(mu.HPDI[1,] ~ weight, data = m4.3.pred, 
           col = "navy", type = "l", add = TRUE)
plotPoints(mu.HPDI[2,] ~ weight, data = m4.3.pred, 
           col = "navy", type = "l", add = TRUE)

# draw PI region for simulated heights
# shade(height.PI, weight.seq)
plotPoints(height.PI[1, ] ~ weight, data = m4.3.pred, 
           col = "red", type = "l", add = TRUE)
plotPoints(height.PI[2, ] ~ weight, data = m4.3.pred, 
           col = "red", type = "l", add = TRUE)

```

## R code 4.62

```{r, chunk4.62, opts.label = "original"}
sim.height <- sim(m4.3, data = list(weight = weight.seq), n = 1e4)
height.PI <- apply(sim.height, 2, PI, prob = 0.89)
```

```{r, chunk4.62a, opts.label = "updated"}
sim.height <- sim(m4.3, data = m4.3.pred, n = 1e4)
height.PI <- apply(sim.height, 2, PI, prob = 0.89)
```


## R code 4.63

This is roughly how `sim()` works.

```{r, chunk4.63}
post <- extract.samples(m4.3)
weight.seq <- 25:70
sim.height <- sapply(weight.seq, function(weight)
  rnorm(
    n = nrow(post),
    mean = post$a + post$b * weight,
    sd = post$sigma
  ))
height.PI <- apply(sim.height, 2, PI, prob = 0.89)
```

## R code 4.64

```{r, chunk4.64, opts.label = "original"}
library(rethinking)
data(Howell1)
d <- Howell1
str(d)

```

```{r, chunk4.64a, opts.label = "updated"}
library(rethinking)
data(Howell1)
str(Howell1)

```
## R code 4.65

```{r, chunk4.65, opts.label = "original"}
d$weight.s <- (d$weight - mean(d$weight)) / sd(d$weight)

```

```{r, chunk4.65a, opts.label = "updated"}
require(mosaic)  # for zscore()
Howell1 <-
  Howell1 %>%
  mutate(
    weight.s = zscore(weight),
    weight.s2 = weight.s^2
  )
```

## R code 4.66

```{r, chunk4.66, opts.label = "original"}
d$weight.s2 <- d$weight.s ^ 2
m4.5 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1 * weight.s + b2 * weight.s2,
    a ~ dnorm(178, 100),
    b1 ~ dnorm(0, 10),
    b2 ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = d
)
```

```{r, chunk4.66a, opts.label = "updated"}
m4.5 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1 * weight.s + b2 * weight.s2,
    a ~ dnorm(178, 100),
    b1 ~ dnorm(0, 10),
    b2 ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = Howell1
)
```

## R code 4.67

```{r, chunk4.67}
precis(m4.5)

```

## R code 4.68

```{r, chunk4.68, opts.label = "original"}
weight.seq <- seq(from = -2.2,
                  to = 2,
                  length.out = 30)
pred_dat <- list(weight.s = weight.seq, weight.s2 = weight.seq ^ 2)
mu <- link(m4.5, data = pred_dat)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob = 0.89)
sim.height <- sim(m4.5, data = pred_dat)
height.PI <- apply(sim.height, 2, PI, prob = 0.89)

```
```{r, chunk4.68a, opts.label = "updated"}
m4.5.pred <-
  data_frame(
    weight.s = seq(from = -2.2, to = 2, length.out = 30),
    weight.s2 = weight.s^2
  )
mu <- link(m4.5, data = m4.5.pred)
sim.height <- sim(m4.5, data = m4.5.pred)
mu.PI <- apply(mu, 2, PI, prob = 0.89)
m4.5.pred <-
  m4.5.pred %>%
  mutate(
    mu.mean = apply(mu, 2, mean),
    mu.lo = apply(mu, 2, PI)[1,],
    mu.hi = apply(mu, 2, PI)[2,],
    sim.lo = apply(sim.height, 2, PI)[1,],
    sim.hi = apply(sim.height, 2, PI)[2,]
    )
```

## R code 4.69

```{r, chunk4.69, opts.label = "original"}
plot(height ~ weight.s, d, col = col.alpha(rangi2, 0.5))
lines(weight.seq, mu.mean)
shade(mu.PI, weight.seq)
shade(height.PI, weight.seq)
```

```{r, chunk4.69a, opts.label = "updated"}
xyplot(sim.hi + mu.hi + mu.mean + mu.lo + sim.lo ~ weight.s, 
       data = m4.5.pred,  type = "l", auto.key = list(lines = TRUE, points = FALSE))
plotPoints(height ~ weight.s, data = Howell1, col = rangi2, alpha = 0.5, add = TRUE)
```
```{r, chunk4.69b, opts.label = "updated"}
ggplot(aes(x = weight.s), data = m4.5.pred) +
  geom_point(aes(y = height), data = Howell1, color = rangi2) +
  geom_line(aes(y = mu.lo), color = "navy") +
  geom_line(aes(y = mu.hi), color = "navy") +
  geom_line(aes(y = sim.lo), color = "red") +
  geom_line(aes(y = sim.hi), color = "red") 
```

## R code 4.70

```{r, chunk4.70, opts.label = "original"}
d$weight.s3 <- d$weight.s ^ 3
m4.6 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1 * weight.s + b2 * weight.s2 + b3 * weight.s3,
    a ~ dnorm(178, 100),
    b1 ~ dnorm(0, 10),
    b2 ~ dnorm(0, 10),
    b3 ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = d
)
```


```{r, chunk4.70a, opts.label = "updated"}
Howell1 <- 
  Howell1 %>% mutate(weight.s3 = weight.s^3)

m4.6 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1 * weight.s + b2 * weight.s2 + b3 * weight.s3,
    a ~ dnorm(178, 100),
    b1 ~ dnorm(0, 10),
    b2 ~ dnorm(0, 10),
    b3 ~ dnorm(0, 10),
    sigma ~ dunif(0, 50)
  ),
  data = Howell1
)
```

## R code 4.71

```{r, chunk4.71, opts.label = "original"}
plot(height ~ weight.s,
     d,
     col = col.alpha(rangi2, 0.5),
     xaxt = "n")
```

```{r, chunk4.71a, opts.label = "updated"}
xyplot(
  height ~ weight, data = Howell1,
  col = rangi2, alpha = 0.5)
```

## R code 4.72

```{r, chunk4.72, opts.label = "original"}
plot(height ~ weight.s,
     d,
     col = col.alpha(rangi2, 0.5),
     xaxt = "n")
at <- c(-2, -1, 0, 1, 2)
labels <- at * sd(d$weight) + mean(d$weight)
axis(side = 1,
     at = at,
     labels = round(labels, 1))

```

```{r, chunk4.72a, opts.label = "updated"} 
xyplot(height ~ weight, data = Howell1, 
       col = rangi2,
       scales=list(
         x = list(at =  round((-2:2) * sd(Howell1$weight) + mean(Howell1$weight), 1))
       )
)

```
## R code 4.73

```{r, chunk4.73, opts.label = "original"}
plot(height ~ weight, data = Howell1,
     col = col.alpha(rangi2, 0.4))

```

```{r, chunk4.73a, opts.label = "updated"}
xyplot(height ~ weight, data = Howell1,
       col = rangi2, alpha = 0.4)
```

