---
title: "Statistical Rethinking (Code)"
author: "Chapter 5"
date: "February, 2017"
output:
  html_document: 
    fig_height: 3.5
  pdf_document: 
    fig_height: 3.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.show = "hold")
require(mosaic)
require(rethinking)
require(ggformula)
theme_set(theme_minimal())
trellis.par.set(theme = theme.mosaic())
set.seed(123456)
inc.model <- TRUE
inc.lattice <- FALSE
inc.gg <- FALSE
inc.gf <- TRUE
```

Code from *Statistical Rethinking* modified by R Pruim is shown below.  Differences to the oringal include:

  * a preference for putting data into containers (data frames, mostly), rather than working with lose vectors.
  * use of `lattice` and/or `ggplot2` rather than base graphics
  * use of `tidyverse` for data transformation
  * better (in my opinion) naming conventions

### Intro: Waffle Houses and Divorce

The following model suggests that divorce rates are higher
in states with more Waffle Houses (per capita).  

```{r}
data(WaffleDivorce)
WaffleDivorce <- 
  WaffleDivorce %>% 
  mutate(
    WaffleHouses.pm = WaffleHouses / Population
  )
m5.0 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bW * WaffleHouses.pm,
    a ~ dnorm(10, 10),
    bW ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = WaffleDivorce
)
precis(m5.0, digits = 3)
```

Here is a plot.
```{r, include = inc.gf}
require(ggformula)   # need the version from rpruim on github
coef(m5.0)
gf_point(Divorce ~ WaffleHouses.pm, data = WaffleDivorce, color = rangi2) %>%
  gf_coefline(coef = coef(m5.0))
```


### R code 5.1

```{r, chunk5.1, include = inc.model}
# load data
library(rethinking)
data(WaffleDivorce)

# standardize predictor
WaffleDivorce <- 
  WaffleDivorce %>% 
  mutate(MedianAgeMarriage.s = zscore(MedianAgeMarriage))

# fit model
m5.1 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bA * MedianAgeMarriage.s,
    a ~ dnorm(10, 10),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = WaffleDivorce
)
```

### R code 5.2

```{r, chunk5.2}
# compute percentile interval of mean
m5.1.pred <-
  data.frame(
    MedianAgeMarriage.s = seq(from = -3, to = 3, by = 0.25)
  )
m5.1.link <- link(m5.1, data = m5.1.pred)
m5.1.pred <- 
  m5.1.pred %>% 
  mutate(
    mu.PIlo = apply(m5.1.link, 2, PI)[1,],
    mu.PIhi = apply(m5.1.link, 2, PI)[2,]
  )
```

```{r, chunk5.2gf, include = inc.gf}
# plot it all
gf_point(Divorce ~ MedianAgeMarriage.s + label:Loc, color = rangi2, 
         data = WaffleDivorce) %>% 
  gf_ribbon(mu.PIlo + mu.PIhi ~ MedianAgeMarriage.s, data = m5.1.pred,  
            alpha = 0.2) %>%
  gf_coefline(coef = coef(m5.1), col = "red", alpha = 0.5) %>%
plotly::ggplotly()
```

### R code 5.3

```{r, chunk5.3, include = inc.model}
WaffleDivorce <-
  WaffleDivorce %>% 
  mutate(Marriage.s = zscore(Marriage))

m5.2 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bR * Marriage.s,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = WaffleDivorce
)
```

```{r, chunk5.3a}
m5.2.pred <-
  data.frame(
    Marriage.s = seq(from = -3, to = 3, by = 0.25)
  )
m5.2.link <- link(m5.2, data = m5.2.pred)
m5.2.pred <- 
  m5.2.pred %>% 
  mutate(
    mu.PIlo = apply(m5.2.link, 2, PI)[1,],
    mu.PIhi = apply(m5.2.link, 2, PI)[2,]
  )
```

```{r, chunk5.3g, include = inc.gg, include = inc.gg}
# plot it all
ggplot(WaffleDivorce) +
  geom_point(aes(y = Divorce,  x = Marriage.s, state = Loc), col = rangi2) +
  geom_abline(intercept = coef(m5.2)["a"], slope = coef(m5.2)["bR"], 
              col = "red", alpha = 0.5) +
  geom_ribbon(data = m5.2.pred, alpha = 0.1,
              aes(x = Marriage.s, ymin = mu.PIlo, ymax = mu.PIhi))
plotly::ggplotly()
```

```{r, chunk5.3gf, include = inc.gg, include = inc.gf}
# plot it all
gf_point(Divorce ~ Marriage.s, data = WaffleDivorce, col = rangi2) %>%
  gf_coefline(coef = coef(m5.2), col = "red", alpha = 0.5) %>%
  gf_ribbon(mu.PIlo + mu.PIhi ~ Marriage.s, data = m5.2.pred, alpha = 0.1)
plotly::ggplotly()
```

### R code 5.4

```{r, chunk5.4, include = inc.model}
m5.3 <- map(
  alist(
    Divorce ~ dnorm(mu, sigma),
    mu <- a + bR * Marriage.s + bA * MedianAgeMarriage.s,
    a ~ dnorm(10, 10),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = WaffleDivorce
)
precis(m5.3)
```

### R code 5.5

```{r, chunk5.5}
plot(precis(m5.3))
```

### R code 5.6

```{r, chunk5.6, include = inc.model}
m5.4 <- map(
  alist(
    Marriage.s ~ dnorm(mu, sigma),
    mu <- a + b * MedianAgeMarriage.s,
    a ~ dnorm(0, 10),
    b ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = WaffleDivorce
)
```

### R code 5.7 -- computing residuals (the hard way)

```{r, chunk5.7}
# for each state, compute expected value at MAP and residual
WaffleDivorce <-
  WaffleDivorce %>% 
  mutate(
    mu.m5.4 = coef(m5.4)['a'] + coef(m5.4)['b'] * MedianAgeMarriage.s,
    resid.m5.4 = Marriage.s - mu.m5.4
  )
```

### R code 5.7a -- computing residuals (the "easy" way)

`link()` takes care of computing the model value for each posterior sample
so we don't have to access the coefficients directly.  As models become 
more complicated, this will be easier.  Also, it provides us with more 
information, since we have a distribution of model responses for each
case.

We are using the defaults here, but `link()` takes two additional arguments
in addition to the model.

  * `n`: number of posterior samples to use.  (The default is 1000.)  Increase this to reduce sampling variability (at the cost of time).
  * `data`: optional data set to use in place of original data from `map()`.
  Use this to make predictions using any predictor values you like.

```{r, chunk5.7a}
# for each state, compute expected value at MAP and residual
m5.4.link <- link(m5.4) 
WaffleDivorce <-
  WaffleDivorce %>%
  mutate(
    mu = apply(m5.4.link, 2, mean),
    resid = Marriage.s - mu
  )
WaffleDivorce %>% select(Loc, mu, resid) %>% head(3)
```

### R code 5.8

```{r, chunk5.8gf, include = inc.gf, include = inc.gf}
gf_segment(mu.m5.4 + Marriage.s ~ MedianAgeMarriage.s + MedianAgeMarriage.s, 
           data = WaffleDivorce, color = "red", alpha = 0.5) %>%
  gf_coefline(coef = coef(m5.4)) %>%
  gf_point(Marriage.s ~ MedianAgeMarriage.s, col = rangi2, data = WaffleDivorce) 
plotly::ggplotly()
```

### R code 5.9

```{r, chunk5.9}
# prepare new counterfactual data
m5.3.pred <- 
  data_frame(
    Marriage.s = seq(from = -3, to = 3, by = 0.25),
    MedianAgeMarriage.s = mean(WaffleDivorce$MedianAgeMarriage.s)
  )

# compute counterfactual mean divorce (mu)
mu <- link(m5.3, data = m5.3.pred)
# simulate counterfactual divorce outcomes
R.sim <- sim(m5.3, data = m5.3.pred, n = 1e4)

m5.3.pred <-
  m5.3.pred %>% 
  mutate(
     mu.mean = apply(mu, 2, mean),
     mu.PIlo =  apply(mu, 2, PI)[1,],
     mu.PIhi =  apply(mu, 2, PI)[2,],
     R.PIlo = apply(R.sim, 2, PI)[1,],
     R.PIhi = apply(R.sim, 2, PI)[2,]
  )
```


```{r, 5.9gf, include = inc.gg, include = inc.gf}
gf_line(mu.mean ~ Marriage.s, data = m5.3.pred, color = "gray50") %>%
  gf_ribbon(mu.PIlo + mu.PIhi ~ Marriage.s, fill = "gray50", alpha = 0.2) %>%
  gf_ribbon(R.PIlo + R.PIhi ~ Marriage.s, fill = "gray50", alpha = 0.2) %>%
  gf_labs( y = "Divorce rate", caption = "MedianAgeMarriage.s = 0")
```




### R code 5.10

```{r, chunk5.10}
m5.3.pred2 <-
  data_frame(
    MedianAgeMarriage.s =  seq(from = -3, to = 3, by = 0.25),
    Marriage.s = mean(WaffleDivorce$Marriage.s)
    )

mu <- link(m5.3, data = m5.3.pred2, n = 1e4)
A.sim <- sim(m5.3, data = m5.3.pred2, n = 1e4)

m5.3.pred2 <-
  m5.3.pred2 %>%
  mutate(
    mu.mean = apply(mu, 2, mean),
    mu.PIlo = apply(mu, 2, PI)[1,],
    mu.PIhi = apply(mu, 2, PI)[2,],
    A.PIlo = apply(A.sim, 2, PI)[1,],
    A.PIhi = apply(A.sim, 2, PI)[2,]
  )

```
```{r, 5.10gf, include = inc.gf, include = inc.gf}
gf_line(mu.mean ~ MedianAgeMarriage.s, data = m5.3.pred2, color = "red") %>%
  gf_ribbon(mu.PIlo + mu.PIhi ~ MedianAgeMarriage.s, alpha = 0.1) %>%
  gf_ribbon(A.PIlo + A.PIhi ~ MedianAgeMarriage.s, alpha = 0.1) %>%
  gf_labs( y = "Divorce rate", caption = "Marriage.s = 0")
```

### R code 5.11

```{r, chunk5.11}
# call link without specifying new data
# so it uses original data
divorce.mu  <- link(m5.3)
# simulate observations
# again no new data, so uses original data
divorce.sim <- sim(m5.3, n = 1e4)

# summarize samples across cases
m5.3.pred3 <-
  WaffleDivorce %>% 
  mutate(
    mu.mean = apply(divorce.mu, 2, mean),
    mu.PIlo = apply(divorce.mu, 2, PI)[1,],
    mu.PIhi = apply(divorce.mu, 2, PI)[2,],
    divorce.PIlo = apply(divorce.sim, 2, PI)[1,],
    divorce.PIhi = apply(divorce.sim, 2, PI)[2,]
  )
```

### R code 5.12


```{r, chunk5.12gf, include = inc.gf, include = inc.gf}
gf_pointrange(mu.mean + divorce.PIlo + divorce.PIhi ~ Divorce, data = m5.3.pred3,
              color = rangi2) %>%
  gf_abline(slope = 1, intercept = 0) %>%
  gf_labs(x =  "Observed divorce", y = "Predicted divorce")
```

### R code 5.13

```{r, chunk5.13, include = inc.gg || inc.gf}
# make most recent ggplot interactive
# adding text to the geom_point() above makes it available on hover here.
plotly::ggplotly()
```

### R code 5.14

```{r, chunk5.14}
# compute residuals
m5.3.pred3 <-
  m5.3.pred3 %>% 
  mutate(divorce.resid = Divorce - mu.mean,
         state = reorder(Loc, divorce.resid))
```

```{r, chuk5.14g, include = inc.gg}
ggplot(m5.3.pred3) +
  geom_linerange(
    col = "gray50", alpha = 0.5, size = 0.8,
    aes(x = state,
        y = divorce.resid, ymin = Divorce - divorce.PIhi, ymax = Divorce - divorce.PIlo)) +
  geom_linerange(
    col = "gray50", alpha = 0.8, size = 1.2,
    aes(x = state, 
        ymin = Divorce - mu.PIhi, ymax = Divorce - mu.PIlo)) +
  geom_point(aes(x = state, y = divorce.resid)) + 
  coord_flip() 
```

```{r, chuk5.14gf, include = inc.gf, include = inc.gf, fig.height = 5}
gf_linerange((Divorce - divorce.PIlo) + (Divorce - divorce.PIhi) ~ Loc, 
             data = m5.3.pred3, alpha = 0.2, size = 0.9) %>%
  gf_linerange((Divorce - mu.PIlo) + (Divorce - mu.PIhi) ~ Loc, 
             data = m5.3.pred3, alpha = 0.2, size = 1.4) %>%
  gf_point(divorce.resid ~ Loc, data = m5.3.pred3) + 
  coord_flip() 
```

```{r, inlcude = inc.gf}
WaffleDivorce <-
  WaffleDivorce %>%
  mutate(
    WaffleHouses.pc = WaffleHouses / Population * 1e6,
    divorce.resid = Divorce - apply(link(m5.3), 2, mean) 
  )
gf_point(divorce.resid ~ WaffleHouses.pc + label:Loc, 
         data = WaffleDivorce) %>%
  gf_smooth(divorce.resid ~ WaffleHouses.pc, method = "lm", size = 0.5) %>%
  plotly::ggplotly()
```
### R code 5.15

```{r, chunk5.15}
n <- 100
Sim5.15Data <-
  data_frame( 
    x_real = rnorm(n),         # x_real as Gaussian with mean 0 and stddev 1
    x_spur = rnorm(n, x_real), # x_spur as Gaussian with mean = x_real
    y =  rnorm(n, x_real)      # y as Gaussian with mean=x_real
  )

m_both <-
  map(
    alist(
      y ~ dnorm(mu, sigma),
      mu <- a + b_real * x_real + b_spur * x_spur,
      a ~ dnorm(0, 3),
      b_real ~ dnorm(0, 3),
      b_spur ~ dnorm(0, 3),
      sigma ~ dlnorm(0, 3)),
    data = Sim5.15Data
  ) 

m_real <-
  map(
    alist(
      y ~ dnorm(mu, sigma),
      mu <- a + b_real * x_real, 
      a ~ dnorm(0, 3),
      b_real ~ dnorm(0, 3),
      sigma ~ dlnorm(0, 3)),
    data = Sim5.15Data
  ) 

m_spur <- 
  map(
    alist(
      y ~ dnorm(mu, sigma),
      mu <- a + b_spur * x_spur,
      a ~ dnorm(0, 3),
      b_spur ~ dnorm(0, 3),
      sigma ~ dlnorm(0, 3)),
    data = Sim5.15Data
  )
lapply(list(both = m_both, real = m_real, spur = m_spur), precis)
```

```{r, chunk5.15g, include = inc.gg}
ggplot(Sim5.15Data, aes(x = x_real, y = y)) +
  geom_point() +
  geom_abline(intercept = coef(m_real)["a"], slope = coef(m_real)["b_real"])

ggplot(Sim5.15Data, aes(x = x_spur, y = y)) +
  geom_point() +
  geom_abline(intercept = coef(m_spur)["a"], slope = coef(m_spur)["b_spur"]) 
```
```{r, chunk5.15gf, include = inc.gf}
gf_point(y ~ x_real, data = Sim5.15Data) %>%
  gf_coefline(coef = coef(m_real))

gf_point(y ~ x_spur, data = Sim5.15Data) %>%
  gf_coefline(coef = coef(m_spur))
```

## The milk data set (masked relationships)

### Source

Comparative primate milk composition data, from Table 2 of 
Hinde and Milligan. 2011. **Evolutionary Anthropology** 20:9-23.

### Variables

  * `clade`: Broad taxonomic group

  * `species`: Species name

  * `kcal.per.g`: Kilocalories per gram of milk

  * `perc.fat`: Percent fat

  * `perc.protein`: Percent protein

  * `perc.lactose`: Percent lactose

  * `mass`: Body mass of mother, in kilograms

  * `neocortex.perc`: Percent of brain mass that is neocortex

### R code 5.16

```{r, chunk5.16}
data(milk)
```

### R code 5.17

```{r, chunk5.17, error = TRUE, include = inc.model}
# This fails.
m5.5 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = milk
)
```

### R code 5.18

```{r, chunk5.18}
# Here's why the previous chunk failed -- missing data!
favstats( ~ neocortex.perc, data = milk)
```

### R code 5.19

```{r, chunk5.19}
# Let's grab just the rows that have no missing data.
MilkCC <- milk %>% filter(complete.cases(.))
favstats(~ neocortex.perc, data = MilkCC)
```

### R code 5.20

```{r, chunk5.20, include = inc.model}
m5.5 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = MilkCC
)
```

### R code 5.21

```{r, chunk5.21}
precis(m5.5, digits = 3)
```

### R code 5.22

```{r, chunk5.22}
coef(m5.5)["bn"] * (76 - 55)
```

### R code 5.23

```{r, chunk5.23}
pred.data <- data.frame(neocortex.perc = 50:80)
mu <- link(m5.5, data = pred.data, n = 1e4)
m5.5.pred <-
  pred.data %>%
  mutate(
    mu.mean = apply(mu, 2, mean),
    mu.PIlo = apply(mu, 2, PI)[1,],
    mu.PIhi = apply(mu, 2, PI)[2,]
  )
```

```{r, chunk5.23gf, include = inc.gf}
gf_point(kcal.per.g ~ neocortex.perc + label:species, data = MilkCC,
         color = rangi2) %>%
  gf_line(mu.mean ~ neocortex.perc, data = m5.5.pred) %>%
  gf_ribbon(mu.PIlo + mu.PIhi ~ neocortex.perc, data = m5.5.pred,
            alpha = 0.2) %>%
plotly::ggplotly()
```
### R code 5.24

```{r, chunk5.24}
MilkCC <- 
  MilkCC %>% 
  mutate(log.mass = log(mass))
```

### R code 5.25

```{r, chunk5.25, include = inc.model}
m5.6 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bm * log.mass,
    a ~ dnorm(0, 100),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = MilkCC
)
precis(m5.6)
```

### R code 5.26

```{r, chunk5.26, include = inc.model}
m5.7 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bn * neocortex.perc + bm * log.mass,
    a ~ dnorm(0, 100),
    bn ~ dnorm(0, 1),
    bm ~ dnorm(0, 1),
    sigma ~ dunif(0, 1)
  ),
  data = MilkCC,
  start = list(a = 0, bn = 0, bm = 0, sigma = 0.50)
)
precis(m5.7)
```

### R code 5.27

```{r, chunk5.27}
m5.7.pred <-
  data_frame(
    neocortex.perc = 50:80,
    log.mass = mean(log(MilkCC$mass))
  )
m5.7.link <- link(m5.7, data = m5.7.pred, n = 1e4)
m5.7.pred <-
  m5.7.pred %>% 
  mutate(
    mu.mean = apply(mu, 2, mean),
    mu.PIlo = apply(mu, 2, PI)[1,],
    mu.PIhi = apply(mu, 2, PI)[2,]
  )
```

```{r, chunk5.27gf, include = inc.gf}
gf_point(kcal.per.g ~ neocortex.perc + label:species, data = MilkCC) %>%
gf_ribbon(mu.PIlo + mu.PIhi ~ neocortex.perc, data = m5.7.pred, alpha = 0.2) %>%
gf_line(mu.mean ~ neocortex.perc, data = m5.7.pred, alpha = 0.5) %>%
plotly::ggplotly()
``` 

### R code 5.28

```{r, chunk5.28}
# simulating a masking relationship
# n = number of cases
# rho = correlation between two variables x_pos and x_neg
sim_masking <- function(n = 100, rho = 0.7) {
  data_frame(
    x_pos = rnorm(n),
    x_neg = rnorm(n, rho * x_pos, sd = sqrt(1 - rho^2)),
    y = rnorm(n, x_pos - x_neg, sd = 1)   # y equally associated to each var
  )
}
MaskingData <- sim_masking()
splom(MaskingData)        # splom = scatter plot matrix (lattice)
GGally::ggpairs(MaskingData)   # ggplot2 version
```

## More isn't always merrier

So far we have seen examples where adding in additional variables
helps us understand what is going on.  But adding variables to the model 
can actually make thigs worse in some situations.

We start with a simulated example and then return to our milk data set.

### R code 5.29

```{r, chunk5.29}
sim_legs <- function(n = 100) {
  data_frame(
    height = rnorm(n, 10, 2),    # units = ??
    leg_prop = runif(n, 0.4, 0.5),
    leg_left = leg_prop * height + rnorm(n, 0, 0.2),
    leg_right = leg_prop * height + rnorm(n, 0, 0.2)
  )
}
LegHeight <- sim_legs()
```

### R code 5.30

```{r, chunk5.30, include = inc.model}
m5.8 <- map(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + bl * leg_left + br * leg_right,
    a ~ dnorm(10, 100),
    bl ~ dnorm(2, 10),
    br ~ dnorm(2, 10),
    sigma ~ dunif(0, 10)
  ),
  data = LegHeight
)
precis(m5.8)
```

### R code 5.31

```{r, chunk5.31}
GGally::ggpairs(LegHeight)
plot(precis(m5.8))

```

### R code 5.32

```{r, chunk5.32}
m5.8.post <- extract.samples(m5.8)
```

```{r, chunk5.32gf, include = inc.gf}
gf_point(bl ~ br, data = m5.8.post, col = rangi2, alpha = 0.1, pch = 16)
```

### R code 5.33

```{r, chunk5.33gf, include = inc.gf}
gf_dens( ~ (bl + br), data = m5.8.post) 
```

### R code 5.34

```{r, chunk5.34, include = inc.model}
m5.9 <- map(alist(
  height ~ dnorm(mu, sigma),
  mu <- a + bl * leg_left,
  a ~ dnorm(10, 100),
  bl ~ dnorm(2, 10),
  sigma ~ dunif(0, 10)
),
data = LegHeight)
precis(m5.9)
plot(precis(m5.9))
```

### R code 5.35

```{r, chunk5.35}
library(rethinking)
data(milk)
```

### R code 5.36

```{r, chunk5.36, include = inc.model}
# kcal.per.g regressed on perc.fat
m5.10 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bf * perc.fat,
    a ~ dnorm(0.6, 10),
    bf ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = milk
)

# kcal.per.g regressed on perc.lactose
m5.11 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bl * perc.lactose,
    a ~ dnorm(0.6, 10),
    bl ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = milk
)

lapply(list(fat = m5.10, lactose = m5.11), precis, digits = 3)
```

### R code 5.37

```{r, chunk5.37, include = inc.model}
m5.12 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + bf * perc.fat + bl * perc.lactose,
    a ~ dnorm(0.6, 10),
    bf ~ dnorm(0, 1),
    bl ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = milk
)
lapply(list(fat = m5.10, lactose = m5.11, `fat + lactose` = m5.12), precis, digits = 3)
```

### R code 5.38

```{r, chunk5.38gg, include = inc.gg || inc.gf}
GGally::ggpairs(milk %>% select(kcal.per.g, perc.fat, perc.lactose))
```

### R code 5.39

```{r, chunk5.39}
cor(milk$perc.fat, milk$perc.lactose)
# alternative syntax using mosaic package
cor(perc.fat ~ perc.lactose, data = milk)
```

### R code 5.40

```{r, chunk5.40}
library(rethinking)
data(milk)
collinearity_sim <- 
  function(rho = 0.9, data = milk) {
    data <- 
      data %>% 
      mutate(
        x = rnorm(nrow(data), mean = rho * perc.fat,
               sd = sqrt((1 - rho^2) * var(perc.fat)))
      )
    model = lm(kcal.per.g ~ perc.fat + x, data = data)
    sqrt(diag(vcov(model)))[2] # stddev of parameter
  } 
collinearity_sim <-
  Vectorize(collinearity_sim, "rho")

SimData5.40 <-
  expand.grid(
    r = seq(from = 0, to = 0.99, by = 0.01),
    rep = 1:100) %>%
  mutate(
    sd = collinearity_sim(r)
  ) 
```


```{r, chunk5.40gf, include = inc.gf} 
gf_point(sd ~ r, data = SimData5.40, alpha = 0.01) %>%
  gf_spline(sd ~ r)
```

### R code 5.41

```{r, chunk5.41}
# simulate data where growth is inhibited by fungus, which is inhibit by soil treatments
# number of plants
SimData5.13 <-
  expand.grid(
    treatment = c(0, 1),
    rep = 1:50              # 50 plants in each treatment group
  ) %>% 
  mutate(
    height0 = rnorm(100, 10, 2),  # initial heights of plants
    # fungus grows in half of control group and 10% of treatment group
    fungus = rbinom(100, size = 1, prob = 0.5 - 0.4 * treatment),
    # mean growth is 5 without fungus, 2 with fungus
    height1 = height0 + rnorm(100, 5 - 3 * fungus)
  )
```

### R code 5.42

If we use `treatment` and `fungus` to predict growth, `treatment` appears to have no effect.
But we know it does (since we simulated it that way).  The impact of `treatment` is 
masked by the use of `fungus`, since the way `treatment` affects growth is by inhibiting
fungus.  

```{r, chunk5.42, include = inc.model}
m5.13 <- map(
  alist(
    height1 ~ dnorm(mu, sigma),
    mu <- a + bh * height0 + bt * treatment + bf * fungus,
    a ~ dnorm(0, 100),
    c(bh, bt, bf) ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = SimData5.13
)
precis(m5.13)
```

### R code 5.43

If we remove `fungus`, we can see the effect of `treatment`.

```{r, chunk5.43, include = inc.model}
m5.14 <- map(
  alist(
    height1 ~ dnorm(mu, sigma),
    mu <- a + bh * height0 + bt * treatment,
    a ~ dnorm(0, 100),
    c(bh, bt) ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
  ),
  data = SimData5.13,
  start = list(a = 2, bh = 1, bt = 1, sigma = 1)
)
precis(m5.14)
```

## Categorical Predictors -- 2 levels


### R code 5.44

```{r, chunk5.44}
data(Howell1)
```

### R code 5.45

```{r, chunk5.45, include = inc.model}
m5.15 <- 
  map(
    alist(
      height ~ dnorm(mu, sigma),
      mu <- a + bm * male,
      a ~ dnorm(178, 100),
      bm ~ dnorm(0, 10),
      sigma ~ dunif(0, 50)
    ),
    data = Howell1)
precis(m5.15)
```

### R code 5.46

```{r, chunk5.46}
m5.15.post <- 
  extract.samples(m5.15) %>% 
  mutate(
    mu.male = a + bm
  )
PI(m5.15.post$mu.male)
```

### R code 5.47

```{r, chunk5.47, iclude = inc.model}
m5.15b <- 
  map(
    alist(
      height ~ dnorm(mu, sigma),
      mu <- af * (1 - male) + am * male,
      af ~ dnorm(178, 100),
      am ~ dnorm(178, 100),
      sigma ~ dunif(0, 50)
    ),
    data = Howell1,
    start = list(af = 178, am = 178, sigma = 3)
  )
```

## Categorical Predictors -- More than 2 Levels

Coding a $k$-level categorical variable as $1, 2, 3, \dots k$, is not a good
idea in most cases.  (Think about what $\mu_i = a + b x_i$ would say in 
that case -- it's usually not the model we want.)  Instead, we need a 
different encoding that uses *multiple predictors*.

### R code 5.48

```{r, chunk5.48}
data(milk)
tally(~ clade, data = milk)
```

### R code 5.49 - 50

Here we create variables that are indicators for a particular level
of `clade`.  That is, for each clade $c$ we create a new variable
$x_c$ defined as

$$
x_c = [\![\mathrm{clade} = c  ]\!] 
$$

```{r, chunk5.49}
milk <-
  milk %>% 
  mutate(
    clade.Ape = ifelse(clade == "Ape", 1, 0),
    clade.NWM = ifelse(clade == "New World Monkey", 1, 0),
    clade.OWM = ifelse(clade == "Old World Monkey", 1, 0),
    clade.S = ifelse(clade == "Strepsirrhine", 1, 0)
  )
milk %>% select(species, matches("clade")) %>% sample_n(5)
```

### R code 5.51

```{r, chunk5.51, include = inc.model}
m5.16 <- map(
  alist(
    kcal.per.g ~ dnorm(mu, sigma),
    mu <- a + b.NWM * clade.NWM + b.OWM * clade.OWM + b.S * clade.S,
    a ~ dnorm(0.6, 10),
    b.NWM ~ dnorm(0, 1),
    b.OWM ~ dnorm(0, 1),
    b.S ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = milk
)
precis(m5.16)
```

### R code 5.52

```{r, chunk5.52}
# sample posterior
m5.16.post <- 
  extract.samples(m5.16) %>%
  mutate(
    # compute averages for each category
    mu.ape = a,
    mu.NWM = a + b.NWM,
    mu.OWM = a + b.OWM,
    mu.S   = a + b.S
  )

# summarize using precis 
# computes mean, sd and PI for each variable in data frame
precis(m5.16.post, digits = 3)
```

### R code 5.53

```{r, chunk5.53}
quantile( ~ (mu.NWM - mu.OWM), data = m5.16.post, probs = c(0.025, 0.5, 0.975))
```

### R code 5.54

`map()` provides ways to deal with all of the clade indicator variables systematically.
First, we convert to the numbers $1, 2, 3, 4$.

```{r, chunk5.54}
milk <-
  milk %>% 
  mutate(
    clade_id = coerce_index(clade)
  )
milk %>% select(matches("clade")) %>% sample_n(5)
```

### R code 5.55

Now we fit a model with four coefficients `a[1], a[2], a[3], a[4]`.

```{r, chunk5.55, include = inc.model}
m5.16_alt <- 
  map(
    alist(
      kcal.per.g ~ dnorm(mu, sigma),
      mu <- a[clade_id],
      a[clade_id] ~ dnorm(0.6, 10),
      sigma ~ dunif(0, 10)
    ),
    data = milk)
precis(m5.16_alt, depth = 2)
```

## lm() and map()

Linear models can be fit in a non-Bayesian way using `lm()`.  Because `lm()` knows the model
is linear, it uses terser notation for the model -- you just list the variables and `lm()` supplies
the coefficients to multiply by each one (and an intercept, unless you remove it).  Also
`lm()` knows how to convert categorical data into multiple predictors, so you only have 
list the categorical variable and `lm()` takes care of the rest.

### R code 5.56

```{r, chunk5.56, eval = FALSE}
m5.17 <- lm(y ~ 1 + x, data = d)
m5.18 <- lm(y ~ 1 + x + z + w, data = d)
```

### R code 5.57

```{r, chunk5.57, eval = FALSE}
m5.19a <- lm(y ~ 1 + x, data = d)
m5.19b <- lm(y ~ x, data = d)
```

### R code 5.58

```{r, chunk5.58, eval = FALSE}
m5.20 <- lm(y ~ 0 + x, data = d)
m5.21 <- lm(y ~ x - 1, data = d)
```

### R code 5.59

```{r, chunk5.59, eval = FALSE}
m5.22 <- lm(y ~ 1 + as.factor(season), data = d)
```

### R code 5.60

```{r, chunk5.60, eval = FALSE}
m5.23 <- lm(y ~ 1 + x + x2 + x3, 
            data = d %>% mutate(x2 = x^2, x3 = x^3))
```

### R code 5.61

```{r, chunk5.61, eval = FALSE}
m5.24 <- lm(y ~ 1 + x + I(x^2) + I(x^3), data = d)
```

### R code 5.62

If you know how to make a model using `lm()`, `glimmer()` will help you create
a formula list for `map()` that does the same job.

```{r, chunk5.62}
data(cars)
glimmer(dist ~ speed, data = cars)
```

```{r, chunk5.62a}
data(KidsFeet, package = "mosaicData")
glimmer(length ~ width + sex, data = KidsFeet)
# this shows how lm is coding the categorical variable
KidsFeet %>% head(5) %>% select(name, sex, length, width)
model.matrix(length ~ width + sex, data = KidsFeet) %>% head(5)
```

As you might guess, folks interested in a Bayesian approach to linar models (and some generalizations
of linear models) have also written functions to make those models easier to describe and it.
The `rstanarm` and `brms` packages are two such packages.  Here's a quick example using the 
default priors.

```{r}
require(rstanarm)
model <- stan_lm(length ~ width + sex, data = KidsFeet, prior = NULL)
model
```