---
title: "Birds"
author: "Stat 241 -- Spring 2017"
date: ''
output:
  pdf_document:
    fig_height: 2.5
    fig_width: 5.5
  html_document:
    fig_height: 2.5
    fig_width: 5.5
  word_document:
    fig_height: 2.5
    fig_width: 3.5
---

```{r, setup, include=FALSE}
# Load packages here 
require(rstan)
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  cache = TRUE,
  tidy = FALSE,     # display code as typed
  size = "small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_bw())
```

#### The Data 

```{r}
data(birdextinct, package = "LearnBayes")
head(birdextinct, 3)
```

Measurements on breedings pairs of landbird species were collected from 16 islands about Britain over several decades.

 * `species` -- name of bird species
 * `time` -- average time of extinction on the islands
 * `nesting` -- average number of nesting pairs
 * `size` -- size of the species: 1 or 0 if large or small
 * `status` -- status of the species: 1 or 0 if resident or migrant

Let's do a bit of renaming to make it easier to remember the coding.
```{r}
Birds <- birdextinct %>%
  mutate(large = size, resident = status, 
         size = ifelse(large, "large", "small"),
         status = ifelse(resident, "resident", "migratory"),
         combo = paste(size, status),
         logtime = log(time))  
```

#### A model

```{r, mBirds, include = FALSE}
mBirds <- 
  map2stan(
    alist(
      logtime ~ dnorm( mu , sigma ),
      mu <- Intercept + b_nesting * nesting + 
            b_large * large + b_resident * resident,
      Intercept ~ dnorm(0,10),
      b_nesting ~ dnorm(0,10),
      b_large ~ dnorm(0,10),
      b_resident ~ dnorm(0,10),
      sigma ~ dcauchy(0,2)
    ),
    data = Birds)
```

```{r, ref.label = "mBirds", eval = FALSE, cache = FALSE}
```

```{r}
precis(mBirds)
```

#### Model Predictions

  1. What does this model predict for the mean log extinction time for a species 
that currently has 4 nesting pairs? That depends, of course, on the covariates 
(four combinations of size and status). Compute these four estimates several 
ways:

    * "by hand" from the output above
    * using `link()`
    * using `extract.samples()` but not `link()`

  2. How is what `link()` does different from what you did "by hand"? Does it
  matter? Why or why not? Is the answer the same for all types of models?
 
  3. How is `sim()` different from `link()`?  Explain how to obtain
  the output from `sim()` using the output from `extract.samples()`.
  

```{r, include = FALSE}
Birds_pred <-
  expand.grid( large = 0:1, resident = 0:1 ) %>% mutate(nesting = 4)
Birds_link <- link(mBirds, data = Birds_pred)
Birds_sim  <- sim(mBirds, data = Birds_pred)
Birds_pred %>%
  mutate(
    link.mean = apply(Birds_link, 2, mean),
    link.lo = apply(Birds_link, 2, PI)[1, ],
    link.hi = apply(Birds_link, 2, PI)[2, ],
    sim.mean = apply(Birds_sim, 2, mean),
    sim.lo = apply(Birds_sim, 2, PI)[1, ],
    lsimhi = apply(Birds_sim, 2, PI)[2, ])
```

  

```{r, include = FALSE}
Birds_link <- link(mBirds)
Birds_sim  <- sim(mBirds)
Birds2 <-
  Birds %>% 
  mutate(
    link.mean = apply(Birds_link, 2, mean),
    link.lo = apply(Birds_link, 2, PI)[1, ],
    link.hi = apply(Birds_link, 2, PI)[2, ],
    sim.mean = apply(Birds_sim, 2, mean),
    sim.lo = apply(Birds_sim, 2, PI)[1, ],
    sim.hi = apply(Birds_sim, 2, PI)[2, ])
```

  4. How do you create this plot and what does it tell us about the model?

  
```{r, echo = FALSE}
gf_point(logtime ~ nesting, data = Birds2, alpha = 0.5, shape = 1) %>%
  gf_line( link.mean ~ nesting, data = Birds2, alpha = 0.5, color = "red") %>%
  gf_linerange( link.lo + link.hi ~ nesting, data = Birds2, size = 0.9, alpha = 0.5) %>%
  gf_linerange( sim.lo + sim.hi ~ nesting, data = Birds2, size = 0.5, alpha = 0.5) %>%
  gf_text(logtime ~ nesting + label:species, data = Birds2, color = "transparent") %>% 
  gf_text(logtime ~ nesting + label:species + color:combo, 
          data = Birds2 %>% filter(logtime < sim.lo | logtime > sim.hi),
          size = 2) %>%
  gf_facet_grid(status ~ size) 
  # plotly::ggplotly()
```

  5. Do you like this plot better or worse?  Why?
  
```{r, echo = FALSE}
gf_point(logtime ~ nesting + color:combo, data = Birds2, alpha = 0.5) %>%
  gf_line( link.mean ~ nesting + color:combo + group:combo, data = Birds2, alpha = 0.5) %>%
  gf_linerange( link.lo + link.hi ~ nesting + color:combo, 
                data = Birds2, size = 0.9, alpha = 0.5) %>%
  gf_linerange( sim.lo + sim.hi ~ nesting + color:combo, 
                data = Birds2, size = 0.3, alpha = 0.5) %>%
  gf_text(logtime ~ nesting + label:species + color:combo, 
          data = Birds2 %>% filter(logtime < sim.lo | logtime > sim.hi),
          size = 2) 
  # plotly::ggplotly()
```
 
  6. What would it mean to add **interaction** to this model?  How 
  might the plots above change if we used a model with interaction?
  
  7. (Describe how to) fit this model using the grid method.
  
  8. Why did we abandon the grid method in favor of `map()` and then `map2stan()`?