---
title: "Reed Frogs"
author: "Stat 341 — Spring 2017"
date: "April, 2017"
output:
  pdf_document: 
    fig_height: 2.0
    fig_width: 4.0
  html_document: default
---

```{r setup, include=FALSE}
require(rstan)
require(rethinking)
require(mosaic)
require(ggformula)
theme_set(theme_bw())
knitr::opts_chunk$set(
  fig.show = "hold",
  cache = TRUE,
  echo = TRUE,
  eval = TRUE)
```

<!-- ## Reed Frogs -->

<!-- Tadpoles really, but frogs is easier to type. -->
<!-- This example is about survival rates of tadpoles in various tanks. -->
<!-- We are gong to start with some very simple models.  (You will get -->
<!-- to explore some more interesting models for homework.) -->

```{r, chunk12.1, include = FALSE}
Frogs <- rethinking::reedfrogs %>% 
  mutate(tank = 1:n())   # add the tank cluster variable
glimpse(Frogs)
```

### Models

```{r, chunk12.2a, include = TRUE, results = "hide", message = FALSE, warning = FALSE}
# complete pooling
m12.0 <- 
  map2stan(
    alist(
      surv ~ dbinom(density, p),
      logit(p) <- a,
      a ~ dnorm(0, 5)
    ), data = Frogs, refresh = 0)
```

```{r, chunk12.2, include = TRUE, results = "hide", message = FALSE, warning = FALSE}
# no pooling
m12.1 <- 
  map2stan(
    alist(
      surv ~ dbinom(density, p),
      logit(p) <- a_tank[tank],
      a_tank[tank] ~ dnorm(0, 5)
    ), data = Frogs, refresh = 0)
```

```{r, chunk12.3, eval = FALSE}
# "Varying Intercepts model" (a.k.a., "partial pooling")
m12.2 <- 
  map2stan(
    alist(
      surv ~ dbinom(density, p),
      logit(p) <- a_tank[tank],
      a_tank[tank] ~ dnorm(a, sigma),
      a ~ dnorm(0, 1),
      sigma ~ dcauchy(0, 1)
    ), data = Frogs, iter = 4000, chains = 4, refresh = 0 )
```

```{r, chunk12.3a, include = FALSE}
# "Varying Intercepts model"
m12.2 <- 
  map2stan(
    alist(
      surv ~ dbinom(density, p),
      logit(p) <- a_tank[tank],
      a_tank[tank] ~ dnorm(a, sigma),
      a ~ dnorm(0, 1),
      sigma ~ dcauchy(0, 1)
    ),
    data = Frogs, iter = 4000, chains = 4, refresh = 0 )
```

  1. How many parameters does each model have? What does each paramter "mean"?

### Comparing models

```{r, chunk12.4}
compare(m12.0, m12.1, m12.2)
```


  2. How do you explain the effective number of parameters vs. the acutal
  number of paramters for each model?
  
\newpage

#### A plot

The plot below shows the survival proportion in each tank (`m12.1`) and the 
model predicted survival proportion (`m12.2`).  The horizontal line is the
median survival proportion over all the tanks.

```{r, chunk12.5, include = FALSE}
# extract Stan samples
m12.2_post <- extract.samples(m12.2)

# compute median intercept for each tank and transform with logistic()
Frogs <- Frogs %>%
  mutate(
    propsurv.est = logistic(apply(m12.2_post$a_tank, 2, median))
  )
```

```{r, chunk12.5a, echo = FALSE}
gf_point(propsurv ~ tank + shape::"m12.1" + color::"m12.1", data = Frogs) %>%
  gf_point(propsurv.est ~ tank + shape::"m12.2" + color::"m12.2", data = Frogs) %>% 
  # mark posterior median probability across tanks
  gf_hline(yintercept = logistic(median(m12.2_post$a)), lty = 2, alpha = 0.5) %>%
  gf_facet_grid( ~ density, scale="free") %>%
  gf_labs(y = "proportion surviving") %>%
  gf_refine(scale_shape_manual(values = c(16, 1)))
```

  3. How do the two survival proportion compare in each tank?  Why is there this pattern?
  
  4. This phenomonon is sometimes called "shrinkage".  Why does is it called that?
  
  5. Which tanks exhibit the most shrinkage?  Why do you think this is?  (Note: there are 
  at least two components to this answer.)
  
  6. What good and/or bad features about the model does this plot reveal or suggest?
  

```{r, chunk12.5b, include = FALSE}
gf_point(propsurv ~ tank + shape::"m12.1" + color::"m12.1", data = Frogs) %>%
  gf_point(propsurv.est ~ tank + shape::"m12.2" + color::"m12.2", data = Frogs) %>% 
  # mark posterior median probability across tanks
  gf_hline(yintercept = logistic(median(m12.2_post$a)), lty = 2, alpha = 0.5) %>%
  gf_facet_grid(pred ~ density, scale="free") %>%
  gf_labs(y = "proportion surviving") %>%
  gf_refine(scale_shape_manual(values = c(16, 1)))
```
```{r, chunk12.5c, include = FALSE}
gf_point(propsurv ~ tank + shape::"m12.1" + color::"m12.1", data = Frogs) %>%
  gf_point(propsurv.est ~ tank + shape::"m12.2" + color::"m12.2", data = Frogs) %>% 
  # mark posterior median probability across tanks
  gf_hline(yintercept = logistic(median(m12.2_post$a)), lty = 2, alpha = 0.5) %>%
  gf_facet_grid(size ~ density, scale="free") %>%
  gf_labs(y = "proportion surviving") %>%
  gf_refine(scale_shape_manual(values = c(16, 1)))
```

### The distribution of survival across tanks

`m12.2` models a population of tanks.  

  7. What does the model say about the population of tanks?
  8. How can we look at what the posterior distribution says about this?
  

```{r, chunk12.6, include = FALSE}
# show first 100 populations in the posterior
D <- data_frame(x = c(-5, 6), y = c(0, 0.4))
p <-  gf_point(y ~ x, data = D, color = "transparent")

for (i in 1:100) {
  p <- 
    p %>% 
    gf_function(fun = dnorm, 
                args = list(mean = m12.2_post$a[i], sd = m12.2_post$sigma[i]),
                alpha = 0.1, color = "navy")
}
p %>%
  gf_labs(x = "log-odds of survival (100 posterior samples)")
```



```{r, include = FALSE}
# sample 8000 imaginary tanks from the posterior distribution
sim_tanks <- rnorm(8000, m12.2_post$a, m12.2_post$sigma)

gf_dens( ~ sim_tanks) %>%
  gf_labs(x = "log-odds of survival (sample of 8000 simulated tanks)")

# transform to probability and visualize
gf_dens( ~ (logistic(sim_tanks))) %>%
  gf_labs(x = "probability survival (sample of 8000 simulated tanks)")
```


## A simulation to illustrate the effects of pooling

To distinguish our simulations from the real data, we refer to ponds rather than tanks.

Three levels of pooling:

  * complete: All ponds are identical.
      * esimate 1 overall survival rate
  * none: Each pond tells you only about itself and not about otehr ponds.
      * estimate separate survival rate for each pond
  * partial: Each pond tell you something about itself and something about all ponds.
      * estimate survival rate for each pond "in context of population of ponds"
      * adaptive regularization (let degree of similarity among ponds drive amount
      of regularization)

Since Bayesian models are **generative**, for any choice of parameters, we should be
able to generate data that matches the way the model thinks data arise.  


  9. For model `m12.2`, what do we need to choose to simulate data?
  
\newpage

### One way to do the simulation

```{r, chunk12.9}
FrogSim <-
  function(
    reps = 15L,
    n = c(5L, 10L, 25L, 35L),
    a = 1.4,
    sigma = 1.5
    ) {
    nponds <- reps * length(n)
    expand.grid(rep = 1:reps, n = n) %>%
      mutate(
        pond = 1:nponds,
        a_true = rnorm(nponds, mean = a, sd = sigma),
        s = rbinom(nponds, prob = logistic(a_true), size = n)
      )
  }
```

  10. Explain what each line of `FrogSim()` is doing.

### Looking at our simulation

Since we know "truth" in our simulation, we can compare the model predictions to "truth".
(We don't get to do this will real data, that's the joy of simulation.)

```{r, chunk12.13, include = FALSE}
SFrogs <- FrogSim()
m12.3 <- map2stan(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a_pond[pond],
    a_pond[pond] ~ dnorm(a, sigma),
    a ~ dnorm(0, 1),
    sigma ~ dcauchy(0, 1)
  ), data = SFrogs, iter = 1e4, warmup = 1000, refresh = 0 )
```

```{r, chunk12.13a, eval = FALSE}
SFrogs <- FrogSim()
m12.3 <- map2stan(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a_pond[pond],
    a_pond[pond] ~ dnorm(a, sigma),
    a ~ dnorm(0, 1),
    sigma ~ dcauchy(0, 1)
  ), data = SFrogs, iter = 1e4, warmup = 1000, refresh = 0 )
```


```{r, chunk12.14, include = FALSE}
precis(m12.3, depth = 2)
ggplot(precis(m12.3, depth = 2))
```

#### Plotting with a function

By combining this code into a function, we can apply it again later without retyping.  This

  * makes it clearer what is going on, and 
  * makes it easier to make systematic changes.

```{r, chunk12.15}
frog_plot <- function(model, data = model@data) {
  data <-
    data %>% as.data.frame() %>% 
    mutate(
      a_pond_est = as.numeric(coef(model)[1:60]),
      p_est = logistic(a_pond_est),
      p_true = logistic(a_true),
      p_raw = s / n,
      nopool_error = abs(p_raw - p_true),
      partpool_error = abs(p_est - p_true)
    )
  gf_point(nopool_error ~ pond, data = data, shape = 16) %>%
    gf_point(partpool_error ~ pond, data = data, shape = 1) %>%
    gf_line(partpool_error ~ pond, 
            data = data %>% group_by(n) %>%
              mutate(partpool_error = mean(partpool_error)),
            linetype = "dashed", color = "navy") %>%
    gf_line(nopool_error ~ pond, 
            data = data %>% group_by(n) %>%
              mutate(nopool_error = mean(nopool_error)),
            linetype = "dotted", color = "red") %>%
    gf_facet_grid( ~ n, scale = "free") %>%
    gf_labs(y = "absolute error")
}
```

Generating a plot from a model is a one-liner now:

```{r}
frog_plot(model = m12.3)
```

  11. Explain what each line of `frog_plot()` is doing.
  12. What does the resulting plot tell us about our three models.
  
#### A second simulation

```{r, chunk12.19, eval = FALSE}
NewData <- FrogSim()
m12.3new <- 
  map2stan(
    m12.3,
    data = NewData,
    iter = 1e4, warmup = 1000, refresh = 0)
frog_plot(m12.3new)
```

```{r, chunk12.19a, include = FALSE}
NewData <- FrogSim()
m12.3new <- 
  map2stan(
    m12.3,
    data = NewData,
    iter = 1e4, warmup = 1000, refresh = 0)
```

```{r chunk12.19b, echo = FALSE}
frog_plot(m12.3new)
```

