---
title: "Test 2 -- Take Home Problems"
author: "Your Name Here"
date: "Stat 341 — Spring 2017"
output:
  html_document:
    fig_height: 2.5
    fig_width: 3.5
  pdf_document:
    fig_height: 2.5
    fig_width: 3.5
  word_document:
    fig_height: 2.5
    fig_width: 3.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  size = "small"
  )
options(width = 95)
```

```{r, include = FALSE}
require(mosaic)
require(rstan)
require(rethinking)
require(ggformula)
theme_set(theme_bw())
```

```{r, loading-data}
# load some data for this test.
load(file = url("http://www.calvin.edu/~rpruim/data/s341/Test2.Rda"))
```


  1. Which model fits the data (in `Pallets2`) better?  How do you know?

  2. Which model would you expect to fit next week's data better?  How much better?
  Explain.
  
  3. If you stopped here (we're not going to do that), what would you conclude
  about overall employee-to-employee variation?  Explain.
  
  4. Which employee does model `m2` think is best (most efficient)?  Worst?
  
  5. How sure is model `m2` that the most efficient employee is more efficient
  than the least efficient employee?  Use posterior sampling to quantify your answer.
  

  6. What does `m3` say about day to day variation in the number of pallets repaired?  
  
  7. Comparing `m2` and `m3`, which seems to have a greater impact on the number 
  of pallets repaired: the employee or the day?  Explain.
  

  8. Why is it important to include both employee and day in our model when
  we are really only interested in differences among the employees?


  9. The only difference between the definitions `m6` and `m7` is the priors.
  What does this difference in priors do?
  
  10. What would happen if we used `dnorm(125, 20)` for both (sets of) 
  normal priors?  Why is this not as good as `m6` and `m7`?
  
  11. What would happen if we used `dnorm(0, 10)` for both (sets of) 
  normal priors?  Why is this not as good as `m6` and `m7`?

  12. Time to pick your favorite model.  Explain how you chose it and provide 
  at least one plot to show how well/poorly it works.
  
  13. Are there any indications of problems with the `map2stan()` fit for your 
  favorite model?  (This should probably have been asked earlier, but I didn't
  want you to have to do it for all of the models.)
  

