---
title: "Problem Set 6 Notes"
author: "R Pruim"
date: "February 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(rethinking)
require(mosaic)
require(ggformula)
```


# Problem Set 6 (chapter 4)

## R Markdown things

Your use of R Markdown is getting better quickly as you figure out how to do the things you want (mostly).  Here are some things to watch out for:

  1. Use the mosaic template to get started.
  
    I created this to give a good boiler plate to start from. You aren't required to use it
    (you might modify it a bit and save a copy with things just the way you like, with your
    name, etc.), but there are some things that will be essentially the same in each document.
    Why type them every time?
  
  1. Avoid overly long lines in code.
  
    Code does not wrap because the alignment of the code helps express its
    organization. Use your linebreaks to make the code more readable and make sure
    nothings spills off the page on the right side.
  
  2. Use autoindent.
  
    If you highlight some code and hit control/command-i, R Studio will autoindent your code.
    This is a good way to make sure you didn't omit a comma or a paren or a curly brace or ...
    It also makes your code more readable.
  
  3. Some items need to be separated by blank lines.
  
    Make sure you have a blank line between R code and the following text and before and after 
    your section headers (`### ` etc. -- also don't forget the space after `###`).
  
  4. Knit from time to time while you are working.
  
    That way if there is an issue, you will know roughly where it is and won't have 
    a surprise and a headache on your hands at the end.  "Restart R and run all chunks" is
    also a good option and nearly the same as knitting.
  
  5. Got math?
  
    See [https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html](https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html) for some notes on how to include 
    mathematical notation.
  
  6. Cheatsheets
  
    RStudio produces a number of cheat sheets, including one on R Markdown.  You can find it at
    [https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/).
    Several of those cheat sheets will be useful for this class.
  
  7. Suppressing things.
  
    Sometimes you don't want all of the R output to show up in your document.  The template 
    includes a chunk near the top that is executed but neither its code nor the results of 
    executing the code are dispalyed.  This is done by adding the chunk option `include = FALSE`.

    Got too many plots?  Try `fig.keep = "last"` to display only the last figure.  Or something like
    `fig.keep = c(1, 4, 7)` to keep the first, fourth, and seventh.
  
     I didn't actually run that code.  I used `eval = FALSE` to tell R not to evaluate the 
     code in that chunk.  That can be useful to show what form a code chunk should have even
     though you don't have the data to execute it.
  
  8. R Chunk options.
  
     There are lots of other code chunk options as well.  Hit tab inside the code chunk to 
     see a list.  Most are pretty self-explanatory.  `fig.width` and `fig.height` can be used
     to modify the height or width of a figure if it needs more or less space, for example.
     
## Deciding what to include in your report/homework

It's hard to give general advice that will work for all situations, but here are some rough 
guidelines.

  1. If you comment on it, display it.
  
    Did you talk about the coefficients of the model?  Make sure they appear in your R output.
    Do you talk about a plot?  Be sure to include the plot.  
  
    For homework, be sure to include enough so that if you make an error, I have some 
    chance of figuring out where it might be.
  
  2. If you don't talk about it, you might not need to include it.
  
    In a normal report, it is generally the case that things that aren't part of
    your "story" should be included in the  output either.  Homework is a little
    different, of course.  If a problem asks for something, you should be sure 
    to include it.  Often, you should comment on it (but not always, I guess).

## Building and interpreting models

  1. In the linear model contact, the intercept and "slope" parameters are often called
  **coefficients**.  The $\sigma$ parameter is not a coefficient of the model, but it 
  is still an important parameter (and you should know how to interpret it).
  
  2. Monotone transformations vs quadratic models
  
    Both monotone transformations (`log()`, `sqrt()`, `1/x`, `x^2`, etc.) and quadratic models 
    (`a + b * x + c * x^2`) can produce fitted values that fall along a curve.  But quadratic 
    models have another feature: they either rise and then fall or fall and then rise.  If the
    presence of a maximum or minimum doesn't make sense for our model, then the quadratic model
    probably isn't what you want.  (You might get lucky and the peak or valley might be outside
    the range of your data, so sometimes you can get away with a quadratic model anyway.)  
  
    In the Kung! height models, a monotone transformation (like the `log()` transformation you
    were asked to try) makes more sense than a quadratic model. 
  
    You can combine both ideas and fit models like `a + b log(x) + c * (log(x))^2`.
  
  3. Interpretting the slope in a model like $y = a + b  \log(x)$.
  
    The main question to ask is what happens when `x` changes.  In this case, an additive 
    change is hard to work with but a multiplicative change is much easier.  For example,
    let's consider the difference `y` values when we incrase `x` by 1%.
  
$$
a + b \log(1.01x) - (a + b \log(x)) = a + b \log(1.01) + b \log x - a - b \log x = b \log(1.01)
$$

  Importantly, this number does not depend on $x$.  So the $b$ coefficient tells
  about the rate at which $y$ increases for a percentage increase in $x$.  (You
  can play the same game with any other percentage you like.)
  
  If you use `log2()`, then you can easily express this in terms of what happens
  when $x$ doubles or halves).

  4. `link()` and `sim()`
  
    We've talked about this in class already, but make sure you understand how
    these two functions work, what sort of output they create, and what you can
    do with that output.  You should also be able to do the same task without
    using `link()` or `sim()` but starting directly from `extract.samples()`.
    
  5. Note: You can use the same posterior samples in both `link()` and `sim()`.
  
    This saves a little time and will make sure that you are using consistent 
    samples for both operations.  Simply save the posterior samples and pass
    them along:
    
```{r eval = FALSE}
posterior_samples <- extract.samples(my_model)
model_link <- link(my_model, post = posterior_samples)
model_sim  <-  sim(my_model, post = posterior_samples)
```

## ggformula is open for business

The `ggformula` package is open for business.  I'll likely lean on that heavily
for graphics going forward becuase (a) it is pretty easy to use, (b) it gives us 
the plots we want for the least amount of typing, and (c) I want to bang on the
code some to find any rough edges that need work.

You can find a vignette describing the package at
[https://rawgithub.com/rpruim/Statistical-Rethinking/master/Examples/ggformula.html](https://rawgithub.com/rpruim/Statistical-Rethinking/master/Examples/ggformula.html)
