---
title: "Problem Set 7 Notes"
author: ""
date: "Stat 341 -- Spring 2017"
output:
  html_document:
    fig_height: 2.5
    fig_width: 3.5
  pdf_document:
    fig_height: 2.5
    fig_width: 3.5
  word_document:
    fig_height: 2.5
    fig_width: 3.5
---

```{r, setup, include=FALSE}
# Load packages here 
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_minimal())
```

## R Markdown stuff

  1. Please make reasonably sized plots.
  
    I suggest you use the template in the `rethinking` package as a 
    starting point.  You can always make an ocassional plot larger 
    as needed.
    
  2. Watch your code formatting
  
    In particular, avoid long lines that don't fit on the page.
    
  3. Use code comments correctly
  
    In-code comments are for comments on your code, not for a 
    discussion of the results.  Put discussion in text blocks, 
    not code blocks.
    
  4. Use the space bar
  
    Use a space after commas, mathematical operators (`+`, `*`, etc.),
    and comment symbols (`#`)
  
See also my comments from PS 6.

## Printer issues

Some of your printers don't do well with shading.  Here is a way to turn off shading and run on an outline if you use `gf_ribbon()`

```{r, eval = FALSE}
gf_ribbon(ymin + ymax ~ x, data = ..., color = "navy", fill = NA)
```

Also, note that `ggplotly()` only works in HTML documents.  You will 
want to leave that out of printed work since you can't interact
with printed plots.

## Bad variable name -- my bad

`LDS.prop` in my code isn't actually a proportion, it is a percent.
I should have checked more carefully what I was getting from 
Wikipedia.  A better name would be `LDS.perc`.

## Plotting counterfactual data

### Group sizes are integers

You know this, of course.  But when you fix the value of group size
to the mean group size, that might not be an integer.  Probably best
to choose an integer value.

### Multiple fits on one plot

Here is an example of how to plot multiple counterfactual plots 
on top of each other.

```{r}
fox.map <- map(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + bA * area + bG * groupsize,
    a ~ dnorm(10, 5),
    c(bA, bG) ~ dnorm(0, 5),
    sigma ~ dunif(0, 5)
  ), 
  data = foxes,
)
fox.link <- link(fox.map)
fox.pred1 <-
  data_frame(
    area = 
```
