---
title: 'Some Review Problems '
author: "Stat 341 -- Spring 2017"
date: ''
output:
  pdf_document:
    fig_height: 2.5
    fig_width: 3.5
  html_document:
    fig_height: 2.5
    fig_width: 3.5
  word_document:
    fig_height: 2.5
    fig_width: 3.5
---

```{r, setup, include=FALSE}
# Load packages here 
require(rstan)
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  cache = TRUE,
  tidy=FALSE,     # display code as typed
  size="small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_bw())
```

### 10H1 recap

In problem 10H1, one parameter in particular showed a substantial difference between
the MAP fit and the Stan fit.  Which parameter?  Why that one?  (This is an example of 
a general principle, and a reason to be cautious about using MAP for logistic regression 
models.)

### Two little models went out to play

Consider the following models. (Note: since no prior is given, both are using
uniform priors.)

```{r}
D1 <- data_frame(x = c(1L, 0L, 1L))
m1 <-  map( alist(
  x ~ dbinom(1, p),
  logit(p) ~ a,
  a ~ dnorm(0, 4)
), data = D1)
```
```{r}
D2 <- data_frame(x = 2L, n = 3L)
m2 <- 
  map( alist( 
    x ~ dbinom(n, p),
    logit(p) ~ a,
    a ~ dnorm(0, 4)
  ), data = D2)
```

  1. Here is the precis output for the first model. 
  What value of `p` is associated with mean of the posterior distribution for `a`?
  
```{r}
precis(m1)
```
 
  2. Let $L_1(p)$ and $L_2(p)$ be the likelihood functions for the two models
  for a given value of $p$. For each pair below, without making any
  computations, determine which is larger and explain how you know.
  
      * $L_1(0.5)$ or $L_1(0.6)$?  
      * $L_2(0.5)$ or $L_2(0.6)$?  
      * $L_1(0.6)$ or $L_2(0.6)$?  
  
  3. These models and data sets are simple enough that you should be able to
  compute the likelihood function for each by hand. Let's do it for `p = 0.5`
  and `p = 0.6` (for each model).
  
  4. Determine the values of $L_1(0.5) / L_2(0.5)$ and $L_1(0.6) / L_2(0.6)$.
  
  5. Without running the code, say what the result of the following code chunk would be
  
```{r, eval = FALSE}
exp(logLik(m1)) / exp(logLik(m2))
```

  6. What output will be produced by `precis(m2)`

