---
title: "Chimpanzees, part 2"
author: "Stat 341 — Spring 2017"
date: "April, 2017"
output:
  pdf_document: 
    fig_height: 1.8
    fig_width: 3
  html_document: default
---

```{r setup, include=FALSE}
require(rethinking)
require(mosaic)
require(ggformula)
theme_set(theme_bw())
library(rstan)
knitr::opts_chunk$set(
  fig.show = "hold",
  cache = TRUE,
  echo = TRUE,
  eval = TRUE)
options(width = 95)
set.seed = 1234
```

### The Data

```{r, chunk10.1, eval = TRUE, cache = FALSE}
library(rethinking)
data(chimpanzees)
Chimps <- chimpanzees %>% 
  mutate( 
    recipient = ifelse(is.na(recipient), -1, recipient),  # to avoid missing data 
    combo = paste0(prosoc_left, "/", condition))          # useful for plotting 
Chimps %>% sample(3)  # 3 random rows
```

### Model with main effects, interaction, and varying intercepts

Note: This is not the same as model `m12.4` in the text.

```{r, chunk12.21a, include = FALSE}
m12.4a <- map2stan(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a + a_actor[actor] + 
      bp * prosoc_left + bc * condition +  bpc * prosoc_left * condition,
    a_actor[actor] ~ dnorm(0, sigma_actor),
    c(a, bp, bc, bpc) ~ dnorm(0, 10),    # nifty short cut!
    sigma_actor ~ dcauchy(0, 1)
  ),
  data = Chimps,
  warmup = 1000, iter = 5000, chains = 4, cores = 3, refresh = 0
)
```

```{r, ref.label = "chunk12.21a", eval = FALSE}
```

  
  1. Describe how to modify the code above in a way that removes `a` from the second
  line of the formula list but still fits the "same model".  
  
  2. In what sense are the two models "the same"?  In what ways are they different?

\newpage

#### Posterior samples

```{r, chunk12.22}
m12.4a_post <- extract.samples(m12.4a) 
glimpse(m12.4a_post) 
```

  3. How does the result of `extract.samples()` change now that we have a multi-level
  model?
  
  4. In a varying intercpets model, each actor has its own intercept.
  In this model, what does the intercept for actor 7 represent?
  
  5. In a Bayesian model, there is a posterior distribution for this intercept.
  Write a little R code that plots the posterior distribution of the 
  intercept for actor 7.  

  
```{r, include = FALSE}
m12.4a_postD <- 
  m12.4a_post %>%
  as.data.frame() %>%
  bind_cols(
    sapply(1:7, function(actor)
      m12.4a_post$a + m12.4a_post$a_actor[, actor]) %>%
      as.data.frame() %>%
      setNames(paste0("intercept.", 1:7))
  )
```

```{r}
gf_dens(~intercept.7, data = m12.4a_postD)
gf_dens(~ logistic(intercept.7), data = m12.4a_postD)
```

## A model with multiple clusters

Again, this is not quite the same as in the book. I've added the other main effect.

```{r, chunk12.23a, include = FALSE}
# prep data 
Chimps <-
  Chimps %>% rename(block_id = block)   # name 'block' is reserved by Stan

m12.5a <- map2stan(
  alist(
    pulled_left ~ dbinom(1, p),
    logit(p) <- a + a_actor[actor] + a_block[block_id] +
      bp * prosoc_left + bc * condition + bpc * condition * prosoc_left,
    a_actor[actor] ~ dnorm(0, sigma_actor),
    a_block[block_id] ~ dnorm(0, sigma_block),
    c(a, bp, bc, bpc) ~ dnorm(0, 10),             # short cut again
    c(sigma_actor, sigma_block) ~ dcauchy(0, 1)   # and again
  ),
  data = Chimps, warmup = 1000, iter = 6000, chains = 4, cores = 3, refresh = 0
)
```

```{r, ref.label = "chunk12.23a", eval = FALSE}
````

  6. Explain how `m12.5a` differs from `m12.4a`.
  
```{r, chunk12.24, fig.height = 4, fig.width = 5, warning =FALSE}
coef(m12.5a)
# precis(m12.5a, depth = 2)     # saving some paper
plot(precis(m12.5a, depth = 2)) # just show the plot
```

  7. What do you noitice when you compare actors to blocks?
  
  8. How do the comparisions below fit into this story?
  
```{r, chunk12.26}
compare(m12.4a, m12.5a)@output %>% 
  mutate(
    logLik = c(logLik(m12.4a), m12.5a = logLik(m12.5a)),
    dlogLik = logLik - logLik(m12.4a))
```


### Posterior samples

Here's a quick trick for turning the list produced by `extract.samples()` into
a data frame you can use for plotting.  Noice the way the names change.

```{r, chunk12.25}
m12.5a_post <- extract.samples(m12.5a) %>% data.frame()
names(m12.5a_post)
```

```{r}
gf_dens(~sigma_block + color::"block_id", data = m12.5a_post) %>%
  gf_dens(~sigma_actor + color::"actor", data = m12.5a_post) %>%
  gf_labs(x = "sigma") %>%
  gf_lims(x = c(0,5))
```

#### Converting to a "long" format.

If we convert our posterior samples from a "wide" format (one row per sample) to
a "long" format (one row for each sample of each parameter, we can plot
posterior distributions for multiple parameters even more easily.

```{r, chunk12.25a, fig.width = 5}
m12.5a_post_long <-  m12.5a_post %>% 
          tidyr::gather(param, value) 
m12.5a_post_long %>% head(3)
gf_dens(~ value + color::param, alpha = 0.6,
        data = m12.5a_post_long %>% filter(grepl("sigma", param)))
```


\newpage

```{r, fig.height = 6}
gf_dens(~ value, 
        data = m12.5a_post_long %>% filter(grepl("a_actor\\.", param))) %>%
  gf_vline(xintercept = 0, color = "red", alpha = 0.5, linetype = "dashed") %>%
  gf_facet_grid( param ~ .)
gf_dens(~ value,
        data = m12.5a_post_long %>% filter(grepl("a_block\\.", param))) %>%
  gf_vline(xintercept = 0, color = "red", alpha = 0.5, linetype = "dashed") %>%
  gf_facet_grid( param ~ .)
```


\newpage

#### R Code 12.27

```{r, chunk12.27}
m12.4a_pred <-
  expand.grid(
    prosoc_left = 0:1,
    condition = 0:1,
    actor = 1:7) %>%
  mutate(
    combo = paste0(prosoc_left, "/", condition)
  )
  
link.m12.4a <- link(m12.4a, data = m12.4a_pred)
m12.4a_pred <- m12.4a_pred %>% 
  mutate(
    p.pred = apply(link.m12.4a, 2, mean),
    p.link.lo = apply(link.m12.4a, 2, PI)[1,],
    p.link.hi = apply(link.m12.4a, 2, PI)[2,]
  )
```

```{r}
gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = m12.4a_pred) %>%
  gf_line(p.pred ~ combo + group::"1", data = m12.4a_pred) %>%
  gf_facet_wrap( ~ actor)
```

#### R Code 12.28

```{r, chunk12.28}
m12.4a_post <- extract.samples(m12.4a) 
m12.4a_postD <- m12.4a_post %>% data.frame()
str(m12.4a_post)
str(m12.4a_postD)
```

#### R Code 12.29

```{r, chunk12.29}
gf_dens( ~ a_actor.5, data = m12.4a_postD)
m12.4a_post_long <- 
  m12.4a_postD %>%
  tidyr::gather(param, value)
gf_dens( ~ value, data = m12.4a_post_long %>% filter(grepl("actor\\.", param))) %>%
  gf_vline(xintercept = 0, color = "red", alpha = 0.5, linetype = "dashed") %>%
  gf_facet_grid( param ~ .)
```

### Creating the link function manually

For some models, it is necessary to create posterior values manually rather than with 
`link()` or `sim()`.  Although this is not required here, we illustrate the manual approach
by way of comparison.

#### R Code 12.30

```{r, chunk12.30}
p.link <- function(prosoc_left, condition, actor, post = m12.4a_post) {
  logodds <- 
    with(post,
         a + a_actor[, actor] + (bp + bpc * condition) * prosoc_left)
  return(logistic(logodds))
} 
str(p.link(0, 0, 2))
```

#### R Code 12.31

```{r, chunk12.31}
G <- expand.grid(
  prosoc_left = 0:1,
  condition = 0:1,
  actor = 1:7) %>%
  mutate(
    combo = paste0(prosoc_left, "/", condition)
  )

manual_link <-
  with(G,    # saves having to type G$ repeatedly
       mapply(p.link, prosoc_left, condition, actor)
  ) 
# Note: a column here for each row in G
str(manual_link)
dim(G)

manual12.4_pred <- 
  G %>%
  mutate(
    p.pred = apply(manual_link, 2, mean),
    p.link.lo = apply(manual_link, 2, PI)[1,],
    p.link.hi = apply(manual_link, 2, PI)[2,]
  )
```

```{r}
# replication of plot from above using home spun link
gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = manual12.4_pred) %>%
  gf_line(p.pred ~ combo + group::"1", data = m12.4a_pred) %>%
  gf_facet_wrap( ~ actor)
```


#### R Code 12.32

```{r, chunk12.32}
# don't need multiple actors this time
G <- expand.grid(
  prosoc_left = 0:1,
  condition = 0:1,
  actor = 1) %>%
  mutate(
    combo = paste0(prosoc_left, "/", condition)
  )
```

#### R Code 12.33

```{r, chunk12.33}
# replace varying intercept samples with zeros
# 1000 samples by 7 actors
a_actor_zeros <- matrix(0, 1000, 7)
```

#### R Code 12.34

```{r, chunk12.34}
# note use of replace list
m12.4a_link2 <- 
  link(
    m12.4a,
    n = 1000,
    data = G,
    replace = list(a_actor = a_actor_zeros)
  )

m12.4a_pred2 <- 
  G %>% 
  mutate(
    p.pred = apply(m12.4a_link2, 2, mean),
    p.link.lo = apply(m12.4a_link2, 2, PI, prob = 0.8)[1,],
    p.link.hi = apply(m12.4a_link2, 2, PI, prob = 0.8)[2,]
  )


gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = m12.4a_pred2) %>%
  gf_line(p.pred ~ combo + group::"doesn't matter what this is", data = m12.4a_pred2) %>%
  gf_labs(x = "prosoc_left/condition", y = "proportion pulled left", title = "average actor")
```

#### R Code 12.35

```{r, chunk12.35}
# replace varying intercept samples with simulations
m12.4a_post <- extract.samples(m12.4a)
a_actor_sims <- 
  rnorm(7000, 0, m12.4a_post$sigma_actor) %>%
  matrix(1000, 7)   # reshape into a 1000 x 7 matrix
```

#### R Code 12.36

```{r, chunk12.36}
m12.4a_link3 <- 
  link(
    m12.4a,
    n = 1000,
    data = G,
    replace = list(a_actor = a_actor_sims)
  )
```
```{r}
m12.4a_pred3 <- 
  G %>% 
  mutate(
    p.pred = apply(m12.4a_link3, 2, mean),
    p.link.lo = apply(m12.4a_link3, 2, PI, prob = 0.8)[1,],
    p.link.hi = apply(m12.4a_link3, 2, PI, prob = 0.8)[2,]
  )


gf_ribbon(p.link.lo + p.link.hi ~ combo + group::"1", data = m12.4a_pred3) %>%
  gf_line(p.pred ~ combo + group::"doesn't matter what this is", data = m12.4a_pred3) %>%
  gf_labs(x = "prosoc_left/condition", y = "proportion pulled left", title = "marginal of actor")
```


#### R Code 12.37

```{r, chunk12.37}
m12.4a_post <- extract.samples(m12.4a) %>%
  as.data.frame() %>%
  mutate(
    sim_a_actor = rnorm(16000, 0, sigma_actor)
  )

Actors50 <- 
  expand.grid(
    actor = 1:50,        # 50 simulated actors
    prosoc_left = 0:1,
    condition = 0:1) %>% 
  mutate(
    combo = paste0(prosoc_left, "/", condition),
    logodds = 
      m12.4a_post$a[actor] +
      m12.4a_post$sim_a_actor[actor] +
      (m12.4a_post$bp[actor] + m12.4a_post$bpc[actor] * condition) * prosoc_left,
    p = logistic(logodds)
  )
```

#### R Code 12.38

```{r, chunk12.38}
gf_line(p ~ combo + group::actor, data = Actors50, alpha = 0.3) %>%
  gf_labs(
    x = "prosoc_left/condition",
    y = "proportion pulled left",
    title = "50 simulated actors"
  )
```

