---
title: "Garden of Forking Paths"
author: "Statistical Rethinking, Chapter 2"
date: "January 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(mosaic)
require(rethinking)

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

# What's in the bag?

There are five bags.  Each one has four milk jug lids that are either blue or green.
Each bag has a different number of blue lids (0, 1, 2, 3, or 4).  If we randomly
select a bag and then sample (with replacement) from the bag, what can we say about
which bag we selected?

We won't know for sure which bag it is, but some bags will be more likely than others.
Can we quantify this?

```{r, echo = FALSE}
Garden <-
  expand.grid(
    bag = 0:4,
    a = 1:4,
    b = 1:4,
    c = 1:4) %>% 
  mutate(
    ya = bag + (a - 2.5) / 5,
    yb = ya + (b - 2.5) / 25,
    yc = yb + (c - 2.5) / 125,
    cola = ifelse(a <= bag, "B", "G"),
    colb = ifelse(b <= bag, "B", "G"),
    colc = ifelse(c <= bag, "B", "G"),
    match = (cola == "B") & (colb == "G") & (colc == "B")
  )

draw_garden <- function(pattern = "***", bags = 0:4) {
  range <- c(min(bags) - 0.5, max(bags) + 0.5)
  pattern <- strsplit(pattern, split = "")[[1]] 
  if (length(pattern) != 3) stop("pattern must be of length 3")
  Garden_plus <-
    Garden %>%
    filter( range[1] <= yc & yc <= range[2] ) %>%
    mutate(
      match_a = (cola == pattern[1] | (pattern[1] == "*")),
      match_ab = match_a & (colb == pattern[2] | (pattern[2] == "*")),
      match = match_ab & (colc == pattern[3] | (pattern[3] == "*"))
    ) 
  p <-
    ggplot(Garden_plus) + 
    geom_segment(data = Garden_plus %>% filter(b ==1, c == 1), 
                 aes(x = 0, xend = 1, y = ya, yend = ya, alpha = match_a), 
                 colour = "gray70", size = 0.2) +
    geom_segment(data = Garden_plus %>% filter(c == 1), 
                 aes(x = 1, xend = 2, y = ya, yend = yb, alpha = match_ab), 
                 colour = "gray70", size = 0.2) +
    geom_segment(data = Garden_plus,
                 aes(x = 2, xend = 3, y = yb, yend = yc, alpha = match), 
                 colour = "gray70", size = 0.2) +
    geom_point(data = Garden_plus %>% filter(b == 1, c == 1), 
               aes(x = 1, y = ya, color = cola, alpha = match_a), size = 1.2) +
    geom_point(data = Garden_plus %>% filter(c == 1), 
               aes(x = 2, y = yb, color = colb, alpha = match_ab), size = 0.9) +
    geom_point(aes(x = 3, y = yc, color = colc, alpha = match), size = 0.5) +
    ylim(min(range) , max(range)) + 
    coord_polar(theta = "y") +
    scale_color_manual(values = c("blue", "green")) +
    scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.2)) +
    theme(
      axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      legend.position="none",
      panel.background=element_blank(),
      panel.border=element_blank(),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      plot.background=element_blank())
    p
    # return(invisible(Garden_plus))
}


```


## Survey says...

Well not a survey really, but what do the data say?

Suppose we draw one lid from the bag and it is Green.  Which bag is most likely?
Which is least likely?  

```{r, warning = FALSE, message = FALSE}
draw_garden() + xlim(0, 1.1)
```

## Let's get some more data

Let's draw two more lids (replacing the previously drawn lid 
and mixing the bag thoroughly each time).

```{r, echo = FALSE}
lids <- "GGG"
```

Now our data is: `r lids`

### A little portion of the garden
Let's consider for a moment just the bag that has 1 Blue 
(and 3 Greens).  How many ways are there to get `r lids`?  
Here's a map through the garden of forking paths for this bag.

```{r, warning = FALSE, message = FALSE}
draw_garden(bags = 1)
```

But not all paths lead to `r lids`.  Let's highlight just that ones that do.

```{r, warning = FALSE, message = FALSE}
draw_garden(bags = 1, pattern = lids)
```

Counting up we see there are 3 * 3 * 3 = 27 such ways.

### A view of the whole garden

Here's the whole garden.

```{r, warning = FALSE, message = FALSE}
draw_garden(bags = 0:4)
```

Again, let's highlight just the paths that lead to `r lids`.

```{r, warning = FALSE, message = FALSE}
draw_garden(bags = 0:4, pattern = lids)
```
It's not too hard to count up the number of paths that lead to
`r lids`, but before we do that, let's come up with a bookkeeping
system that keeps us organized.

## Another View

As we collect more data, it is soon going to be impossible to create maps of the 
whole garden and to count up possible ways of navigating the map to get to the data.
We need some improved methods.  The key insight is that we obtain the total number
of ways for a given bag by multiplying the number of ways to do each step.

```{r, echo = FALSE}
ways <- function(lids, bags = 0:4, prior = rep(1, length.out = length(bags))) {
  lids <- strsplit(lids, "")[[1]]
  lids <- ifelse(lids == "B", 0, 1)  # recode using 0's and 1's
  W <- matrix(NA, nrow = 0, ncol = length(lids))
  for (bag in bags) {
    w <- ifelse(lids == 0, bag, 4 - bag)
    W <- rbind(W, w)
  }
  row.names(W) = paste("bag", bags)
  s <- sum(prior * apply(W, 1, prod)) %>% as.vector()
  res <- 
    data_frame(
      bag = bags,
      prior = prior,
      terms = apply(W, 1, function(x) paste(x, collapse = " * ")),
      ways = apply(W, 1, prod) * prior, 
      prob = ways / sum(ways),
      probf = paste0(ways, "/", sum(ways))
    )
  res %>% data.frame()
}
```

```{r}
ways("GGG")
```

### Another bit of data

```{r, include = FALSE}
lids4 <- "GGGB"
```

Let draw another lid.  Now our data are `r lids4`.  We can use the same
bookkeeping to tally up the ways.
```{r}
ways("GGGB")
```

But we don't have to start all over, we could also start from what we knew
when we had only seen `r lids`:
```{r}
ways("GGG")
ways("B", prior = ways("GGG")$ways)    
```

This means

 1. We can compute the total number of ways (or probabilities) without creating all of the paths.
 
 2. We can update to include new data by taking the informationt we had before
 as prior information and and starting from there.
 
 3. We can hanlde (or get a computer to handle for us) more than just a small number 
 of bags (stay tuned).

Also, it is only the relative number of ways that matters.  We can convert
to probability by dividing by the total number of ways.

```{r, include = FALSE}
bayes_prop <- 
  function(data = c(0, 0, 0, 1), p = c(0, 0.25, 0.50, 0.75, 1.0),  
           prior = 1) {
  Grid <- data_frame(p = p) %>%
    mutate(
      prior_paths = prior,
      prior = prior_paths / sum(prior_paths),
      likelihood0 = dbinom(sum(head(data, -1)), size = length(data)-1, prob = p),
      likelihood = dbinom(sum(data), size = length(data), prob = p),
      paths0 = prior_paths * likelihood0 * 4^(length(data) - 1),
      paths = prior_paths * likelihood * 4^length(data),
      posterior0 = paths0 / sum(paths0),
      posterior = paths / sum(paths),
      d = paste(data, collapse = " "),
      n = length(data)
    )
  return(Grid)
  ggplot(Grid) +
    # geom_point(aes(x = p, y = posterior0, color = "prior")) +
    geom_line(aes(x = p, y = posterior0, color = "prior")) +
    # geom_point(aes(x = p, y = posterior, color = "posterior")) +
    geom_line(aes(x = p, y = posterior, color = "posterior")) +
    # geom_point(aes(x = p, y = paths, color = "paths")) +
    # geom_line(aes(x = p, y = paths, color = "paths")) +
    labs(y = "", title = paste(data, collapse = " "))
} 
```

Here is a grphical view of how our probabilities for each bag are updated as we see 
each bit of data:

```{r, echo = FALSE}
G <-
  Reduce(c, c(0, 0, 0, 1, 0, 1, 0, 0, 1), accumulate = TRUE) %>% 
  lapply(bayes_prop, p = seq(0, 1, length.out = 5)) %>%
  bind_rows()

ggplot(G) +
  # geom_line(aes(x = p, y = posterior, color = "posterior"), alpha = 0.5, size = 1.2) +
  geom_point(aes(x = p, y = posterior, color = "posterior"), size = 2) +
  # geom_line(aes(x = p, y = posterior0, color = "prior"), alpha = 0.5, size = 1.2) +
  geom_point(aes(x = p, y = posterior0, color = "prior"), size = 1.2) +
  facet_wrap(~d)
```



