---
title: "National Economies"
author: "Stat 341 -- Spring 2017"
date: ''
output:
  pdf_document:
    fig_height: 3.0
  html_document:
    fig_height: 3.0
---

```{r, setup, include=FALSE}
# Load packages here 
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  cache = TRUE,
  tidy = FALSE,     # display code as typed
  size = "small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_light())
```

# GDP and ruggedness

> Bad geography tends to be related to bad economies outside of Africa, but African economies seem immune to bad geography. (McElreath 211)

How do we fit models to assess this claim?

## The Data

The data used for this are in the `rugged` data set.
Naming a data frame after one of its variables is not a very good idea.
Let's rename the data and 

 * add in log of 2000 GDP, 
 * add `continent`, which is handy for plotting later
 * remove countries that don't have 2000 GDP data (since they won't be used in the analysis below)
 
```{r, chunk7.1}
library(rethinking)
data(rugged)

Nations <-
  rugged %>% 
  mutate(
    log_gdp = log(rgdppc_2000),
    continent = ifelse(cont_africa, "Africa", "Other")
  ) %>% 
  filter(! is.na(log_gdp))
```

```{r, echo = FALSE}
gf_point(log_gdp ~ rugged + color::continent + shape:continent, 
         data = Nations, size = 2, alpha = 0.7) %>%
  gf_refine(scale_shape_manual(values = c(16, 21)))
```

\newpage

## Approach 1: A bad approach

Here's a bad idea that we will improve upon:

  * split the data into two portions (African and non-African countries)
  * fit a model with log GDP as outcome and ruggedness as predictor in each portion
  
**What's bad about this idea?**

```{r, chunk7.1b, include = FALSE}
# split countries into Africa and not-Africa
Nations_A1 <- Nations %>% filter(cont_africa == 1)
Nations_A0 <- Nations %>% filter(cont_africa == 0)
sapply(list(All = Nations, Africa = Nations_A1, Other = Nations_A0), nrow)
```



```{r, chunk7.2, include = FALSE}
# African nations
m7.1 <- 
  map(
    alist(
      log_gdp ~ dnorm(mu, sigma),
      mu <- a + bR * rugged,
      a ~ dnorm(8, 100),
      bR ~ dnorm(0, 1),
      sigma ~ dunif(0, 10)
    ),
    data = Nations_A1)

# non-African nations
m7.2 <- 
  map(
    alist(
      log_gdp ~ dnorm(mu, sigma),
      mu <- a + bR * rugged,
      a ~ dnorm(8, 100),
      bR ~ dnorm(0, 1),
      sigma ~ dunif(0, 10)
    ),
    data = Nations_A0)
```

```{r, chunk7.3, include = FALSE}
m7.3 <- 
  map(
    alist(
      log_gdp ~ dnorm(mu, sigma),
      mu <- a + bR * rugged,
      a ~ dnorm(8, 100),
      bR ~ dnorm(0, 1),
      sigma ~ dunif(0, 10)
    ),
    data = Nations)
```

\vfill

## Approach 2: This also isn't quite right

**What is wrong with this model?**

```{r, chunk7.4}
m7.4 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged + bA * cont_africa,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data = Nations
)
```

```{r, chunk7.5, include = FALSE}
compare(m7.3, m7.4)
```




```{r, chunk7.6, include = FALSE}
Loggdp_predict <-
  expand.grid(
    cont_africa = 0:1,
    rugged = seq(from = 0, to = 8, by = 0.2)
  ) %>%
  mutate(
    continent = ifelse(cont_africa, "Africa", "Other")
  )
m7.4_link <- link(m7.4, data = Loggdp_predict)
  
# add in means and intervals using m7.4
Loggdp_predict <-
  Loggdp_predict %>%
  mutate(
    m7.4.mu = apply(m7.4_link, 2, mean),
    m7.4.lo = apply(m7.4_link, 2, PI, prob = 0.97)[1,],
    m7.4.hi = apply(m7.4_link, 2, PI, prob = 0.97)[2,]
  )
```

```{r, include = FALSE}
gf_line(m7.4.mu ~ rugged + color:continent,
        data = Loggdp_predict) %>%
gf_ribbon(m7.4.lo + m7.4.hi ~ rugged + fill:continent,
        data = Loggdp_predict) %>%
  gf_point(log_gdp ~ rugged + color:continent, data = Nations) %>%
  gf_refine(scale_x_continuous(trans = "sqrt"))
```



```{r, include = FALSE}
m7.4.resid <- resid(m7.4)
gf_point(m7.4.resid ~ rugged + color:continent, data = Nations) %>%
  gf_smooth(m7.4.resid ~ rugged + color:continent, data = Nations, 
            alpha = 0.2) %>%
  gf_refine(scale_x_continuous(trans = "sqrt"))
```

\vfill

## Approach 3: Fixing the problem

**Describe a single model that allows each country group to have its 
own intercept *and slope*.**

  * Do this as many ways as you can think of.
  
  * Are your different ways just descriptions of the same thing or
  are they actually different models?
  
  * Which formulation(s) do you like better?  Why?

\vfill

## Comparing Appraoches

**What things can you do to compare the approaches?**

  * You don't have to actually do them, just make a list of what
    you would do and what you would look for/at.
  * For each one, jot down the names of the R funcitons you would use.
  * Once your list is complete, if you have time, go ahead and do the 
  things on your list.
   
```{r, chunk7.7, include = FALSE}
m7.5 <- 
  map(
    alist(
      log_gdp ~ dnorm(mu, sigma),
      mu <- a + gamma * rugged + bA * cont_africa,
      gamma <- bR + bAR * cont_africa,
      a ~ dnorm(8, 100),
      bA ~ dnorm(0, 1),
      bR ~ dnorm(0, 1),
      bAR ~ dnorm(0, 1),
      sigma ~ dunif(0, 10)
    ),
    data = Nations
  )
```


```{r, chunk7.8, include = FALSE}
compare(m7.3, m7.4, m7.5)
```


```{r, chunk7.9, include = FALSE}
m7.5b <- 
  map(
    alist(
      log_gdp ~ dnorm(mu, sigma),
      mu <- a + bR * rugged + bAR * rugged * cont_africa + bA * cont_africa,
      a ~ dnorm(8, 100),
      bA ~ dnorm(0, 1),
      bR ~ dnorm(0, 1),
      bAR ~ dnorm(0, 1),
      sigma ~ dunif(0, 10)
    ),
    data = Nations
  )
```


```{r, chunk7.10, include = FALSE}
m7.5_link <- link(m7.5, data = Loggdp_predict)
  
# add in means and intervals
Loggdp_predict <-
  Loggdp_predict %>%
  mutate(
    m7.5.mu = apply(m7.5_link, 2, mean),
    m7.5.lo = apply(m7.5_link, 2, PI, prob = 0.97)[1,],
    m7.5.hi = apply(m7.5_link, 2, PI, prob = 0.97)[2,]
  )
```



```{r, include = FALSE}
gf_line(m7.5.mu ~ rugged + color:continent,
        data = Loggdp_predict) %>%
gf_ribbon(m7.5.lo + m7.5.hi ~ rugged + fill:continent,
        data = Loggdp_predict) 
```


```{r, chunk7.11, include = FALSE}
# plot African nations with regression
gf_point(log_gdp ~ rugged + color:continent, 
         data = Nations) %>% 
  gf_line(m7.5.mu ~ rugged + col:continent, data = Loggdp_predict) %>%
  gf_ribbon(m7.5.lo + m7.5.hi ~ rugged + fill:continent, 
            data = Loggdp_predict) %>%
  gf_labs(y = "log GDP year 2000", x = "Terrain Ruggedness Index")  %>%
  gf_facet_grid(~ continent)
```



```{r, chunk7.12, include = FALSE}
precis(m7.5)
```


```{r, chunk7.13, include = FALSE}
m7.5_post <- extract.samples(m7.5) %>%
  mutate(
    gamma.Africa = bR + bAR * 1,
    gamma.notAfrica = bR + bAR * 0
  )
```

```{r, chunk7.14, include = FALSE}
mean(m7.5_post$gamma.Africa)
mean(m7.5_post$gamma.notAfrica)
```

```{r, chunk7.15, include = FALSE}
gf_dens(~gamma.Africa + color::"Africa", data = m7.5_post) %>%
  gf_dens(~gamma.notAfrica + color::"Other", data = m7.5_post) %>%
  gf_labs(x = "gamma")
```


```{r, chunk7.16, include = FALSE}
gf_dens( ~ (gamma.Africa - gamma.notAfrica), data = m7.5_post) 
prop( ~(gamma.Africa < gamma.notAfrica), data = m7.5_post)
```

    
```{r, chunk7.17b, include = FALSE}
mrugged <- median(~rugged, data = Nations); mrugged

Nations <- 
  Nations %>%
  mutate(
    rugged_group = ifelse( rugged > mrugged, "hi", "lo"))
Loggdp_predict <- 
  Loggdp_predict %>%
  mutate(
    rugged_group = ifelse( rugged > mrugged, "hi", "lo"))

gf_ribbon(
  m7.5.lo + m7.5.hi ~ continent + fill:rugged_group + group:rugged_group, 
  data = Loggdp_predict %>% filter(rugged %in% c(0, 6))) %>%
  gf_line(m7.5.mu ~ continent + color:rugged_group + group:rugged_group, 
          data = Loggdp_predict %>% filter(rugged %in% c(0, 6))) %>%
  gf_point( log_gdp ~ continent + color:rugged_group, data = Nations, 
            position = position_dodge(width = 0.2), alpha = 0.5) 
```


```{r, chunk7.17c, include = FALSE}
gf_ribbon(
  m7.5.lo + m7.5.hi ~ continent + fill:rugged_group + group:rugged_group, 
  data = Loggdp_predict %>% filter(rugged %in% c(0, 6))) %>%
  gf_line(m7.5.mu ~ continent + color:rugged_group + group:rugged_group, 
          data = Loggdp_predict %>% filter(rugged %in% c(0, 6))) %>%
  gf_point( log_gdp ~ continent + color:rugged_group, data = Nations, 
            position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.2), 
            alpha = 0.5)
```



```{r, chunk7.17d, include = FALSE}
quantile(Nations$rugged)
gf_ribbon(
  m7.5.lo + m7.5.hi ~ continent + fill:rugged_group + group:rugged_group, 
  data = Loggdp_predict %>% filter(rugged %in% c(0.5, 2))) %>%
  gf_line(m7.5.mu ~ continent + color:rugged_group + group:rugged_group, 
          data = Loggdp_predict %>% filter(rugged %in% c(0.5, 2))) %>%
  gf_point( log_gdp ~ continent + color:rugged_group, data = Nations, 
            position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.2), 
            alpha = 0.5)
gf_ribbon(
  m7.5.lo + m7.5.hi ~ continent + fill:rugged_group + group:rugged_group, 
  data = Loggdp_predict %>% filter(rugged %in% c(0, 6))) %>%
  gf_line(m7.5.mu ~ continent + color:"gray50" + group:rugged_group, 
          data = Loggdp_predict %>% filter(rugged %in% c(0, 6))) %>%
  gf_point( log_gdp ~ continent + color:rugged_group, data = Nations, 
            position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.2), 
            alpha = 0.5) %>%
  gf_refine(
    scale_fill_manual(values = c("orange", "navy"))
  )
              
```

