---
title: "Tulips"
author: "Stat 341 — Spring 2017"
date: ""
output:
  html_document: 
    fig_height: 3.0
  pdf_document: 
    fig_height: 2.5
    fig_width: 3.0
---

```{r, setup, include=FALSE}
# Load packages here 
require(rethinking)
require(mosaic)   
require(ggformula)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  cache = TRUE,
  tidy = FALSE,     # display code as typed
  size = "small",   # slightly smaller font for code
  fig.show = "hold")   # all plots at end of chunk
theme_set(theme_light())
```


### The data

```{r, chunk7.18}
library(rethinking)
data(tulips)
Tulips <- 
  tulips %>%
  mutate(
    shade.c = shade - mean(shade),
    water.c = water - mean(water)
  )
glimpse(Tulips)
```


## Two Models

**How do these models differ?**

  * What might lead you to fit one but not the other?

  * How would you compare them to decide which model is "better" if you fit both?
  

```{r, chunk7.20}
m7.6 <- map(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- a + bW * water + bS * shade,
    c(a, bW, bS) ~ dnorm(0, 100),
    sigma ~ dunif(0, 100)
  ),
  data = Tulips,
  method = "Nelder-Mead", control = list(maxit = 1e4)
)
m7.7 <- map(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- a + bW * water + bS * shade + bWS * water * shade,
    c(a, bW, bS, bWS) ~ dnorm(0, 100),
    sigma ~ dunif(0, 100)
  ),
  data = Tulips,
  method = "Nelder-Mead", control = list(maxit = 1e4)
)
```

## 9 groups, 9 parameters?

In this set-up, there are only 9 possible combinations of 
water and shade.

```{r}
tally(shade ~ water, data = Tulips)
```

**What predictions do our previous two models give for each of the 9 combinations?**

```{r, chunk7.21}
coeftab(m7.6, m7.7)
```

**Design a model that has 9 parameters, one for each combination.**

 * There is more than one way to do this, choose a way that makes 
   the parameters as interpretable as you can.
   
**What are the advantages/disadvantages of such a model?**

**In what sense is this a "model with interaction"?**

  * How can you tell by looking at the paramters whether this 
  is a large or small "interaction effect"?


```{r}
m7.7a <- map(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- bS[shade] + bW[water] + 
      gamma22 * (shade == 2) * (water == 2) +
      gamma23 * (shade == 2) * (water == 3) +
      gamma32 * (shade == 3) * (water == 2) +
      gamma33 * (shade == 3) * (water == 3),
    bS[shade] ~ dnorm(100, 100),
    bW[water] ~ dnorm(100, 100),
    c(gamma22, gamma23, gamma32, gamma33) ~ dnorm(0, 100),
    sigma ~ dunif(0, 100)
  ),
  data = Tulips,
  method = "Nelder-Mead", control = list(maxit = 1e4)
)
coef(m7.7a)
```

### Six Degrees of freedom

A better way to see the strength of the interaction effect
is to fit a model without interaction terms.

**Design a model with 6 parameters that is the previosu model "minus interaction."**

  * Reparameterize if that is more convenient.


```{r}
m7.7b <- map(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- bS[shade] + bW[water],
    bS[shade] ~ dnorm(100, 100),
    bW[water] ~ dnorm(100, 100),
    sigma ~ dunif(0, 100)
  ),
  data = Tulips,
  method = "Nelder-Mead", control = list(maxit = 1e4)
)
coef(m7.7b)
compare(m7.6, m7.7, m7.7a, m7.7b)
```

```{r, chunk7.22, error = TRUE}
compare(m7.6, m7.7)
```


```{r, chunk7.24}
m7.8 <- map(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- a + bW * water.c + bS * shade.c,
    a ~ dnorm(130, 100),
    c(bW, bS) ~ dnorm(0, 100),
    sigma ~ dunif(0, 100)
  ),
  data = Tulips,
  start = list( a = mean(Tulips$blooms), bW = 0, bS = 0, sigma = sd(Tulips$blooms))
)
m7.9 <- map(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- a + bW * water.c + bS * shade.c + bWS * water.c * shade.c,
    a ~ dnorm(130, 100),
    c(bW, bS, bWS) ~ dnorm(0, 100),
    sigma ~ dunif(0, 100)
  ),
  data = Tulips,
  start = list(a = mean(Tulips$blooms), bW = 0, bS = 0, bWS = 0, 
               sigma = sd(Tulips$blooms))
)
coeftab(m7.8, m7.9)
```

**What does each model predict for a flower bed that gets 
two units of sun and two units of water?**


```{r, chunk7.25, include = FALSE}
k <- coef(m7.7); k
(k["a"] + k["bW"] * 2 + k["bS"] * 2 + k["bWS"] * 2 * 2) %>% setNames("mu.S2W2")
```

#### R Code 7.26

```{r, chunk7.26}
k <- coef(m7.9); k
(k["a"] + k["bW"] * 0 + k["bS"] * 0 + k["bWS"] * 0 * 0) %>% setNames("mu.S0W0")
```

#### R Code 7.27

```{r, chunk7.27}
precis(m7.9)
```

#### R Code 7.28

```{r, chunk7.28}
Blooms_predict <- 
  expand.grid(shade.c = -1:1, water.c = -1:1)

m7.8.link <- link(m7.8, data = Blooms_predict)
m7.9.link <- link(m7.9, data = Blooms_predict)

# We can put two models together above and below or side by side.
# Above/below let's us use facets to compare the model predictions.

Blooms_predict <- 
  bind_rows(
    Blooms_predict %>%
      mutate(
        mu = apply(m7.8.link, 2, mean),
        lo = apply(m7.8.link, 2, PI, 0.97)[1,],
        hi = apply(m7.8.link, 2, PI, 0.97)[2,],
        model = "m7.8"
      ),
    Blooms_predict %>%
      mutate(
        mu = apply(m7.9.link, 2, mean),
        lo = apply(m7.9.link, 2, PI, 0.97)[1,],
        hi = apply(m7.9.link, 2, PI, 0.97)[2,],
        model = "m7.9"
      )
  )

gf_ribbon(lo + hi ~ shade.c, data = Blooms_predict, fill = "navy")  %>%
gf_line(mu ~ shade.c, data = Blooms_predict)  %>%
gf_point(blooms ~ shade.c, data = Tulips, alpha = 0.7, color = "navy") %>%
gf_facet_grid(model ~ paste0("water.c = ", water.c))
```

#### R Code 7.29

```{r, chunk7.29}
m7.x <- lm(blooms ~ water + shade + water:shade, data = Tulips)
coef(m7.x)
```

#### R Code 7.30

```{r, chunk7.30}
# short hand notation
m7.x <- lm(blooms ~ water*shade, data = Tulips)
coef(m7.x)
```

#### R Code 7.31

```{r, chunk7.31}
m7.x <- lm(blooms ~ water*shade - shade, data = Tulips)
coef(m7.x)
```

#### R Code 7.32

```{r, chunk7.32}
m7.x <- lm(blooms ~ water*shade*bed, data = Tulips)
coef(m7.x)
```

#### R Code 7.33

```{r, chunk7.33}
x <- z <- w <- 1
colnames(model.matrix( ~ x * z * w))
```

#### R Code 7.34

```{r, chunk7.34}
data(nettle)
Nettle <-
  nettle %>%
  mutate(lang.per.cap = num.lang / k.pop)
head(Nettle, 3)
```
